---
title: "Test of the function gs_power_npe"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
bibliography: "ggsd.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Test of the function gs_power_npe}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(gt)
library(dplyr)
library(tibble)
library(testthat)
library(gsDesign)
#library(gsDesign2)
devtools::load_all()
```


# Test 1: Examples from spec

## Default
The default of `gs_power_npe` is a single analysis with type I error controlled.
```{r}
x1 <- gs_power_npe(theta = 0) %>% filter(Bound == "Upper")
x2 <- gsDesign2:::gs_power_npe_(theta = 0) %>% filter(Bound == "Upper")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  mutate(`Computated from` = c("new version", "old version")) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

## Fixed bound
```{r}
x1 <- gs_power_npe(theta = c(.1, .2, .3),
                   info = (1:3) * 40,
                   upper = gs_b,
                   upar = gsDesign::gsDesign(k = 3,sfu = gsDesign::sfLDOF)$upper$bound,
                   lower = gs_b,
                   lpar = c(-1, 0, 0)) %>% mutate(`Computated from` = "new version")

x2 <- gs_power_npe_(theta = c(.1, .2, .3),
                    info = (1:3) * 40,
                    upper = gs_b,
                    upar = gsDesign::gsDesign(k = 3,sfu = gsDesign::sfLDOF)$upper$bound,
                    lower = gs_b,
                    lpar = c(-1, 0, 0)) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

## Same fixed efficacy bounds, no futility bound (i.e., non-binding bound), null hypothesis
```{r}
x1 <- gs_power_npe(theta = rep(0, 3),
                   info = (1:3) * 40,
                   upar = gsDesign::gsDesign(k = 3,sfu = gsDesign::sfLDOF)$upper$bound,
                   lpar = rep(-Inf, 3)) %>% mutate(`Computated from` = "new version")

x2 <- gs_power_npe_(theta = rep(0, 3),
                    info = (1:3) * 40,
                    upar = gsDesign::gsDesign(k = 3,sfu = gsDesign::sfLDOF)$upper$bound,
                    lpar = rep(-Inf, 3)) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```


## Fixed bound with futility only at analysis 1; efficacy only at analyses 2, 3
```{r}
x1 <- gs_power_npe(theta = c(.1, .2, .3),
                   info = (1:3) * 40,
                   upper = gs_b,
                   upar = c(Inf, 3, 2),
                   lower = gs_b,
                   lpar = c(qnorm(.1), -Inf, -Inf)) %>% mutate(`Computated from` = "new version")

x2 <- gs_power_npe_(theta = c(.1, .2, .3),
                    info = (1:3) * 40,
                    upper = gs_b,
                    upar = c(Inf, 3, 2),
                    lower = gs_b,
                    lpar = c(qnorm(.1), -Inf, -Inf)) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

## Spending function bounds
```{r}
# Lower spending based on non-zero effect
x1 <- gs_power_npe(theta = c(.1, .2, .3),
                   info = (1:3) * 40,
                   upper = gs_spending_bound,
                   upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
                   lower = gs_spending_bound,
                   lpar = list(sf = gsDesign::sfHSD, total_spend = 0.1, param = -1, timing = NULL)) %>% mutate(`Computated from` = "new version")

x2 <- gs_power_npe_(theta = c(.1, .2, .3),
                    info = (1:3) * 40,
                    upper = gs_spending_bound,
                    upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
                    lower = gs_spending_bound,
                    lpar = list(sf = gsDesign::sfHSD, total_spend = 0.1, param = -1, timing = NULL)) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

## Same bounds, but power under different theta
```{r}
x1 <- gs_power_npe(theta = c(.15, .25, .35),
                   info = (1:3) * 40,
                   upper = gs_spending_bound,
                   upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
                   lower = gs_spending_bound,
                   lpar = list(sf = gsDesign::sfHSD, total_spend = 0.1, param = -1, timing = NULL)) %>% mutate(`Computated from` = "new version")

x2 <- gs_power_npe_(theta = c(.15, .25, .35),
                    info = (1:3) * 40,
                    upper = gs_spending_bound,
                    upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
                    lower = gs_spending_bound,
                    lpar = list(sf = gsDesign::sfHSD, total_spend = 0.1, param = -1, timing = NULL)) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

## Two-sided symmetric spend, O'Brien-Fleming spending
Typically, 2-sided bounds are binding
```{r}
x1 <- gs_power_npe(theta = rep(0, 3),
                   info = (1:3) * 40,
                   binding = TRUE,
                   upper = gs_spending_bound,
                   upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
                   lower = gs_spending_bound,
                   lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL)) %>% mutate(`Computated from` = "new version")

x2 <- gs_power_npe_(theta = rep(0, 3), 
                    info = (1:3) * 40,
                    binding = TRUE,
                    upper = gs_spending_bound,
                    upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
                    lower = gs_spending_bound,
                    lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL)) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
x1 %>% 
  union_all(x2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

## Re-use these bounds under alternate hypothesis
Always use binding = TRUE for power calculations
```{r}
xx1 <- gs_power_npe(theta = c(.1, .2, .3), 
                    info = (1:3) * 40,
                    binding = TRUE,
                    upar = (x1 %>% filter(Bound == "Upper"))$Z,
                    lpar = -(x1 %>% filter(Bound == "Upper"))$Z) %>% mutate(`Computated from` = "new version")

xx2 <- gs_power_npe_(theta = c(.1, .2, .3), 
                     info = (1:3) * 40,
                     binding = TRUE,
                     upar = (x1 %>% filter(Bound == "Upper"))$Z,
                     lpar = -(x1 %>% filter(Bound == "Upper"))$Z) %>% mutate(`Computated from` = "old version")
```

```{r, echo=FALSE}
xx1 %>% 
  union_all(xx2) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```





# Test 2: Fixed Design
The power at the following analysis is expected at 0.975.

```{r}
gs_power_npe(theta = 0, 
             upper = gs_b, upar = qnorm(0.025),
             lower = gs_b, lpar = -Inf)
```


# Test 3: `info` != `info0` != `info1`

If one inputs `info` in `upar`
```{r}
x1_a <- gs_power_npe(theta = c(.1, .2, .3), 
                     info = (1:3) * 80, info0 = (1:3) * 90 + 10, info1 = (1:3) * 70 - 5, info_scale = 0,
                     upper = gs_b, upar = gsDesign::gsDesign(k = 3, sfu = gsDesign::sfLDOF)$upper$bound,
                     lower = gs_b, lpar = c(-1, 0, 0)) %>% mutate(`Computated from` = "new version", `Info scale` = 0)
x1_b <- gs_power_npe(theta = c(.1, .2, .3), 
                     info = (1:3) * 80, info0 = (1:3) * 90 + 10, info1 = (1:3) * 70 - 5, info_scale = 1,
                     upper = gs_b, upar = gsDesign::gsDesign(k = 3, sfu = gsDesign::sfLDOF)$upper$bound,
                     lower = gs_b, lpar = c(-1, 0, 0)) %>% mutate(`Computated from` = "new version", `Info scale` = 1)
x1_c <- gs_power_npe(theta = c(.1, .2, .3), 
                     info = (1:3) * 80, info0 = (1:3) * 90 + 10, info1 = (1:3) * 70 - 5, info_scale = 2,
                     upper = gs_b, upar = gsDesign::gsDesign(k = 3, sfu = gsDesign::sfLDOF)$upper$bound,
                     lower = gs_b, lpar = c(-1, 0, 0)) %>% mutate(`Computated from` = "new version", `Info scale` = 2)
x2 <- gs_power_npe_(theta = c(.1, .2, .3), 
                     info = (1:3) * 80, info0 = (1:3) * 90 + 10, info1 = (1:3) * 70 - 5, 
                     upper = gs_b, upar = gsDesign::gsDesign(k = 3, sfu = gsDesign::sfLDOF)$upper$bound,
                     lower = gs_b, lpar = c(-1, 0, 0)) %>% mutate(`Computated from` = "old version")
```

```{r}
x1_a %>% 
  union_all(x1_b) %>% 
  union_all(x1_c) %>% 
  union_all(x2) %>% 
  arrange(Analysis) %>% 
  group_by(Analysis, Bound) %>% 
  gt() %>% 
  tab_style(
    style = list(cell_fill(color = "#d3edeb")),
    locations = cells_body(rows = `Computated from` == "old version"))
```

# Test 3: Developer Tests

## 1-sided test
```{r}
r = 80
x <- gs_power_npe(theta = 0, 
                  info = (1:3) * 400, 
                  binding = FALSE, r = r,
                  upper = gs_b, #gs_spending_bound, 
                  upar = gsDesign(k = 3, test.type = 1, sfu = sfLDOF)$upper$bound,
                  #list(par = list(sf = gsDesign::sfLDOF, param = NULL, total_spend = 0.025)),
                  lower = gs_b, 
                  lpar = rep(-Inf, 3)) %>% filter(Bound == "Upper")

y <- gsDesign2:::gs_power_npe_(theta = 0, 
                  info = (1:3) * 400, 
                  binding = FALSE, r = r,
                  upper = gs_b, #gs_spending_bound, 
                  upar = gsDesign(k = 3, test.type = 1, sfu = sfLDOF)$upper$bound,
                  #list(par = list(sf = gsDesign::sfLDOF, param = NULL, total_spend = 0.025)),
                  lower = gs_b, 
                  lpar = rep(-Inf, 3)) %>% filter(Bound == "Upper")

z <- gsProbability(k = 3, 
                   theta = 0, 
                   n.I = (1:3) * 400, 
                   b = gsDesign(k = 3, test.type = 1, sfu = sfLDOF)$upper$bound, a = rep(-20, 3), r = r)
```

```{r, echo=FALSE}
tibble(`Calculated from` = rep(c("new version", "old version", "gsDesign"), each = 3),
       Analysis = rep(1:3, 3),
       `upper bound` = c(x %>% filter(Bound == "Upper") %>% select(Z) %>% unlist() %>% as.numeric(),
                         y %>% filter(Bound == "Upper") %>% select(Z) %>% unlist() %>% as.numeric(),
                         z$upper$bound),
       `upper prob` = c(x %>% filter(Bound == "Upper") %>% select(Probability) %>% unlist() %>% as.numeric(),
                        y %>% filter(Bound == "Upper") %>% select(Probability) %>%unlist() %>% as.numeric(),
                        cumsum(z$upper$prob))) %>% 
  arrange(Analysis) %>% 
  group_by(Analysis) %>% 
  gt() 
```



# Test 4: Independent Tests

## Expect equal with mvtnorm for efficacy and futility bounds
```{r}
info <- c(40, 100)
r <- info[1] / info[2]

test<-gs_power_npe(theta = 0,
                   info = info,
                   info0 = NULL,
                   binding = FALSE,
                   upper = gs_spending_bound,
                   upar = list(sf = gsDesign::sfLDOF, param = NULL, total_spend = 0.025),
                   lower = gs_spending_bound,
                   lpar = list(sf = gsDesign::sfLDOF, param = NULL, total_spend = 0.02)
)

test1 <- test%>% filter(Bound == "Upper")
test2 <- test%>% filter(Bound == "Lower")

alpha.t <- 0.025
b.ia <- gsDesign::sfLDOF(alpha = alpha.t, t = r)
alpha.ia <- b.ia$spend

Pb <- function(alpha.t, alpha.ia, r, b){
  temp = mvtnorm::pmvnorm(lower = c(-Inf, b), 
                          upper = c(qnorm(1-alpha.ia), Inf),
                          corr = rbind(c(1, sqrt(r)), c(sqrt(r), 1)))
  return(alpha.t - alpha.ia - temp)
}

b <- uniroot(Pb, c(1.96, 4), alpha.t = alpha.t, alpha.ia = alpha.ia, r = r)

pb <- 1- pnorm(b$root)

expect_equal(object = test1$Z, expected = c(qnorm(1-alpha.ia),b$root), tolerance = 0.001)
expect_equal(object = test1$Probability, expected = cumsum(c(b.ia$spend,pb)), tolerance = 0.001)

  beta.t <- 0.02
  a.ia <- gsDesign::sfLDOF(alpha = beta.t, t = r)
  beta.ia <- a.ia$spend

  Pa <- function(beta.t, beta.ia,  r, a){
    temp <- mvtnorm::pmvnorm(lower = c(-Inf, qnorm(beta.ia)),
                             upper = c(a, Inf),
                             corr = rbind(c(1, sqrt(r)), c(sqrt(r), 1)))
    return(beta.t -  beta.ia - temp)
  }

  a <- uniroot(Pa, c(-4, 1.96), beta.t = beta.t, beta.ia = beta.ia, r = r)

  pa <- pnorm(a$root)

  expect_equal(object = test2$Z, expected = c(qnorm(beta.ia), a$root), tolerance = 0.001)
  expect_equal(object = test2$Probability, expected = cumsum(c(a.ia$spend,pa)), tolerance = 0.001)
```

## Expect equal with `gsDesign::gsProbability` outcome for efficacy bounds

```{r}
info <- c(40, 150, 200)

x <- gs_power_npe(theta = .1,
                  info = info, binding = FALSE,
                  upper = gs_spending_bound,
                  upar = list(sf = gsDesign::sfLDOF, param = NULL, total_spend = 0.025),
                  lower = gs_b,
                  lpar = rep(-Inf, 3)) %>% filter(Bound == "Upper")

y <- gs_power_npe(theta = .1,
                  info = info, binding = FALSE,
                  upper = gs_spending_bound,
                  upar = list(sf = gsDesign::sfLDOF, param = NULL, total_spend = 0.025),
                  lower = gs_b,
                  lpar = rep(-Inf, 3)) %>% filter(Bound == "Upper")

z <- gsDesign::gsProbability(k = 3, theta = .1,
                             n.I = info,
                             a = rep(-20, 3),
                             b = gsDesign(k = 3, test.type=1, sfu = sfLDOF, n.I = info)$upper$bound)
```

```{r, echo=FALSE}
tibble(`Calculated from` = rep(c("new version", "old version", "gsDesign"), each = 3),
       Analysis = rep(1:3, 3),
       `upper bound` = c(x %>% filter(Bound == "Upper") %>% select(Z) %>% unlist() %>% as.numeric(),
                         y %>% filter(Bound == "Upper") %>% select(Z) %>% unlist() %>% as.numeric(),
                         z$upper$bound),
       `upper prob` = c(x %>% filter(Bound == "Upper") %>% select(Probability) %>% unlist() %>% as.numeric(),
                        y %>% filter(Bound == "Upper") %>% select(Probability) %>%unlist() %>% as.numeric(),
                        cumsum(z$upper$prob))) %>% 
  arrange(Analysis) %>% 
  group_by(Analysis) %>% 
  gt() 
```


# Test 5: Compare with `gsDesign` under information-based design

Information-based design is useful when testing for a natural parameter $\delta$ (e.g., treatment difference on a relevant scale such as risk difference) where the variance of the estimate of $\delta$ is unknown. 
The basic *canonical form* of represents information-based design, so it is a particularly simple way to check corresponding basic calculations for sample size, bounds and power in `gs_power_npe()` and `gs_design_npe()`.

## Step 1: set the design assumptions

```{r}
k <- 2           # Number of analyses
test.type <- 4
alpha <- 0.025   # 1-sided Type I error
beta <- 0.15     # Type 2 error (1 - targeted power)
astar <- .1
timing <- 0.4    # Timing (information fraction) at interim analyses
sfu <- sfHSD     # Efficacy bound spending function
sfupar <- -1     # Upper bound spending function parameters, if any
sfl <- sfLDPocock# Lower bound spending function, if used (test.type > 2)
sflpar <- 0      # Lower bound spending function parameters, if any
delta <- 0.1     # Natural parameter difference (assumed value - H0 value)
delta1 <- 0.1    # Natural parameter assumed value
delta0 <- 0      # Natural parameter difference under H0
endpoint <- 'info'
n.fix <- 0
```

# References