---
title: "Trial design with spending under NPH"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteIndexEntry{Trial design with spending under NPH}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
#library(gsdmvn)
devtools::load_all()
library(dplyr)
library(tibble)
library(gsDesign)
library(gsDesign2)
library(gt)
```

# Overview

This vignette covers how to implement designs for trials with spending assuming non-proportional hazards.
We are primarily concerned with practical issues of implementation rather than design strategies, but we will not ignore design strategy.

# Scenario for Consideration

Here we set up enrollment, failure and dropout rates along with assumptions for enrollment duration and times of analyses.

In this example, we assume there are 4 analysis (3 interim analysis + 1 final analysis), and they are conducted after 18, 24, 30, 36 months after the trail starts.
```{r}
K <- 4
analysisTimes <- c(18, 24, 30, 36)
```

And we further assume there is not stratum and the enrollment last for 12 months. 
For the first 2 months, second 2 months, third 2 months and the reminder month, the enrollment rate is $8:12:16:24$. 
Please note that $8:12:16:24$ is not the real enrollment rate. 
Instead, it only specifies the enrollment rate ratio between different duration.
```{r}
enrollRates <- tibble(
  Stratum = "All",
  duration = c(2, 2, 2, 6),
  rate = c(8, 12, 16, 24))

enrollRates %>% 
  gt() %>% 
  tab_header(title = "Table of Enrollment")
```

Moreover, we assume the hazard ratio (HR) of the first 3 month is 0.9 and thereafter is 0.6.
We also assume the the survival time follow a piecewise exponential distribution with a median of 8 month for the first 3 months and 14 month thereafter.
```{r}
failRates <- tibble(
  Stratum = "All",
  duration = c(3, 100),
  failRate = log(2) / c(8, 14),
  hr = c(.9, .6),
  dropoutRate = .001)

failRates %>% 
  gt() %>% 
  tab_header(title = "Table of Failure Rate")
```

# Deriving Power for a Given Sample Size

In this section, we discuss how to drive the power, given a known sample size.

First, we calculate the number of events and statistical information (both under H0 and H1) at targeted analysis times.

```{r}
xx <- AHR(enrollRates = enrollRates, 
          failRates = failRates, 
          totalDuration = analysisTimes)

xx %>% gt()
```

Then, we can use `gs_info_ahr()` to calculate (1) the treatment effect (`theta`), (2) AHR, (3) the statistical information (both under H0 and H1) under the targeted number of events.

```{r}
#Events <- ceiling(xx$Events)
yy <- gs_info_ahr(enrollRates = enrollRates, 
                  failRates = failRates, 
                  events = ceiling(xx$Events)) %>% 
  mutate(timing = info0 / max(info0))

yy %>% 
  gt() %>% 
  fmt_number(columns = 2:8, decimals = 4)
```

Finally, we can calculate the power of `yy` by using `gs_power_npe()`. 

```{r}
zz <- gs_power_npe(
  # set the treatment effet
  theta = yy$theta, 
  # set the statistical information under H0 and H1
  info = yy$info, 
  info0 = yy$info0,
  # set the upper bound
  upper = gs_b, 
  upar = gsDesign(k = K, test.type = 2, sfu = sfLDOF, alpha = .025, timing = yy$timing)$upper$bound,
  # set the lower bound
  lower = gs_b,
  lpar = gsDesign(k = K, test.type = 2, sfu = sfLDOF, alpha = .025, timing = yy$timing)$lower$bound)

zz %>% 
  filter(Bound == "Upper") %>%
  select(Analysis, Bound, Z, Probability, IF) %>% 
  gt() %>% 
  fmt_number(columns = 3:5, decimals = 4)
```
From the above table, we find the power is 0.6267.

# Deriving Sample Size for a Given Power

In this section, we discuss how to calculate the sample size for a given power (we take the given power as 0.9 in this section).
And we discuss the calulation into 2 scenario: (1) fixed design and (2) group sequential design.

```{r}
target_power <- 0.9
```

## Fixed Design

If we were using a fixed design, we would approximate the sample size as follows:

```{r}
minx <- ((qnorm(.025) / sqrt(zz$info0[K]) + qnorm(1 - target_power) / sqrt(zz$info[K])) / zz$theta[K])^2
minx
```

If we inflate the enrollment rates by `minx` and use a fixed design, we will see this achieves the targeted power.

```{r}
gs_power_npe(
  theta = yy$theta[K], 
  info = yy$info[K] * minx, 
  info0 = yy$info0[K] * minx,
  upar = qnorm(.975), 
  lpar = -Inf) %>%
  filter(Bound == "Upper") %>%
  select(Probability)
```

## Group Sequential Design

The power for a group sequential design with the same final sample size is a bit lower:

```{r}
gs_power_npe(
  theta = yy$theta, 
  info = yy$info * minx, 
  info0 = yy$info0 * minx,
  upper = gs_spending_bound, 
  lower = gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL)
  ) %>% 
  filter(Bound == "Upper", Analysis == K) %>% 
  select(Probability) %>% 
  gt()
```

If we inflate this a bit we will be overpowered.

```{r}
gs_power_npe(
  theta = yy$theta, 
  info = yy$info * minx * 1.2, 
  info0 = yy$info0 * minx * 1.2,
  upper = gs_spending_bound, 
  lower = gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL)) %>% 
  filter(Bound == "Upper", Analysis == K) %>% 
  select(Probability) %>% 
  gt()
```

Now we use `gs_design_npe()` to inflate the information proportionately to power the trial.

```{r}
gs_design_npe(
  theta = yy$theta, 
  info = yy$info, 
  info0 = yy$info0,
  upper = gs_spending_bound, 
  lower = gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL)) %>% 
  filter(Bound == "Upper", Analysis == K) %>% 
  select(Probability) %>% 
  gt()
```
