---
title: "Group Sequential Design for Time to Event Outcomes: `gsDesign`, `gsDesign2` vs EAST comparison"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    number_sections: false
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Group Sequential Design for Time to Event Outcomes: `gsDesign`, `gsDesign2` vs EAST comparison}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(tidyr)
library(tibble)
library(gsDesign)
library(gsDesign2)
library(kableExtra)
library(targets)
library(flextable)
library(details)
options(knitr.table.format = "html", knitr.kable.NA = '')
```

# Overview

TO ADD

# Interim Analysis for Efficacy

First, we will compare the outputs from `gsDesign`, `gsDesign2` and EAST for a case where interim analyses are planned for efficacy only. 

## Working Example

- We will use a phase 3 study which compares experimental therapy to a standard of care (SOC) in cancer patients. Although the study outlined here is not a real study, it contains features of commonly used phase 3 designs in oncology.

- The study has two primary endpoints: progression free survival (PFS) and overall survival(OS):
  - PFS is has one interim analysis (IA) for efficacy and futility.
  - OS has two IAs for efficacy only.
  
- It is assumed that the initial sample/power calculations in the protocol are done using EAST.  

### Sample Size Assumptions

- PFS HR=0.6 and median PFS of 9.4 months in the control arm
- OS HR=0.65 and median OS of 3 years in the control arm
-	Enrollment period of 24 months.
- Minimum follow-up of 10 and 44 months for PFS and OS, respectively
- Monthly dropout of 0.019 and 0.001 for PFS and OS, respectively
- 1:1 randomization ratio
- 400 total patients

Target number of events:

- PFS: 236 events with IA at 75% information fraction, 95% power at initially allocated $\alpha$ = 0.0125.
- OS: 219 events with IAs at 60% and 80% information fraction respectively, 82% power at initially allocated $\alpha$ = 0.0125.

### Multiplicity Assumption

TO ADD


## OS Analysis: Efficacy Boundaries and Properties

- Lan-DeMets O’Brien-Fleming alpha-spending function was used to construct the boundaries.

- Efficacy boundaries table calculated using EAST is presented below:


```{r, eval=TRUE}
os_east <- tribble(
  ~analysis, ~value, ~eff_125,  ~eff_25,
  
  'IA3: 60%',   'Z',                   -3.02569409632101,    -2.67335178005868,
  'N=400',      'p (1-sided)',          0.00124031566230105,  0.00375487271138121,
  'Events: 131','HR at bound',          0.589363439145686,    0.626790039898953,
  'Month: 38',  'P(Cross) if HR=1',     0.00124031566236733,  0.00375487271160193, 
  '',           'P(Cross) if HR=0.60',  0.287594389776889,    0.417581743680648,
  
  'IA4: 80%',   'Z',                   -2.58884138850925,    -2.28988990769677,
  'N=400',      'p (1-sided)',          0.00481497196501242,  0.0110138497146098,
  'Events: 175','HR at bound',          0.676112343292131,    0.707371838489819,
  'Month: 50',  'P(Cross) if HR=1',     0.005204198,          0.01216244,
  '',           'P(Cross) if HR=0.60',  0.6078709,            0.7185137,
  
  'FA',         'Z',                    -2.29555066802652,   -2.03037092268016,
  'N=400',      'p (1-sided)',           0.0108507930971346,  0.0211594250017298,
  'Events: 219','HR at bound',           0.733272506020993,   0.760028379543686,
  'Month: 66',  'P(Cross) if HR=1',      0.0125,              0.025,
  '',           'P(Cross) if HR=0.60',   0.8207454,           0.8832428 
  
  
)

os_east |> 
  flextable() |>
  autofit() |>
  set_header_labels(
    analysis = 'Analysis',
     value= 'Value',
     eff_125 = "alpha=0.0125",
     eff_25  = "alpha=0.025"
    ) |>
  hline(i = c(5, 10)) |>
  colformat_double(j = c(3, 4), digits = 4)

```


### Calculations using `gsDesign` and `gsDesign2`

- `gsDesign::gsSurv` is used with `test.type = 1` because the OS IAs are for efficacy only.

- It should be noted that EAST returns integer values for the number of events, so that the actual information fraction specified by the user might not __exactly__ match the actual information fraction, calculated by simply dividing the number of events provided by EAST as can be seen below:

```{r,collapse=FALSE}
timing_east <- c(131/219, 175/219)
timing_east
```


- `gsDesign` and `gsDesign2`, on the other hand, provide non-integer values which match __exactly__ the specified information fraction. For a proper comparison between `gsDesign`, `gsDesign2`,  and EAST we will use the exact timing based on the number of events calculated in EAST, i.e., `timing_east` object.

```{r}
#define input parameters:
enroll_dur <- 24
ss <- 400
enroll_rate <- ss/enroll_dur
minfu_os <- 44
fa_time_os <- enroll_dur + minfu_os
failrate_os <- log(2)/36
hr1_os <- 0.65
hr0 <- 1
do_rate_os <- 0.001


os_scen <- tibble(
  alevel = c(0.0125, 0.025),
  blevel = c(1 - os_east$eff_125[length(os_east$eff_125)], 1 - os_east$eff_25[length(os_east$eff_25)])
  )

#map over two values of alpha
os_gs <-
  os_scen |>
  mutate(
    gs_out = purrr::map2(
      alevel, blevel, function(x, y){
        
    out <- gsSurv(
      k = 3,
      timing = timing_east,
      R = enroll_dur, 
      gamma = enroll_rate,
      eta =  do_rate_os, 
      minfup = minfu_os,
      T = fa_time_os, 
      lambdaC = failrate_os,
      hr = hr1_os,
      hr0 = 1,
      beta = y,
      alpha = x,
      sfu = sfLDOF,
      sfupar = NULL, 
      test.type = 1
      )
    out |> gsBoundSummary()
     
  }))

#dataframes for inputs to be used in gsDesign2 below
enrollRates <- tibble(Stratum="All", duration = enroll_dur, rate = enroll_rate)
failRates <- tibble(Stratum="All", 
                    duration=c(fa_time_os), 
                    failRate = failrate_os, 
                    hr = hr1_os, 
                    dropoutRate = do_rate_os)

os_gs2_a125 <- gsDesign2::gs_power_ahr(
  enrollRates = enrollRates,
  failRates = failRates,
  ratio = 1, 
  events = c(131, 175, 219),
  upper = gs_spending_bound, 
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.0125), 
  test_lower = FALSE
) 

os_gs2_a25 <- gsDesign2::gs_power_ahr(
  enrollRates = enrollRates,
  failRates = failRates,
  ratio = 1, 
  events = c(131, 175, 219),
  upper = gs_spending_bound, 
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025), 
  test_lower = FALSE
) 

```

<div style="float: left; width: 50%;">
- Summary from `gsDesign::gsSurv` for $\alpha = 0.0125$:

```{r}
os_gs |>
  dplyr::filter(alevel == 0.0125) |>
  tidyr::unnest(gs_out) |>
  dplyr::select(-c(alevel, blevel)) |>
  flextable::flextable() |>
  flextable::autofit() |>
  flextable::hline(i = c(5, 10))

```
</div>

<div style="float: right; width: 50%;">
- Summary from `gsDesign::gsSurv` for $\alpha = 0.025$:

```{r}
os_gs |>
  dplyr::filter(alevel == 0.025) |>
  tidyr::unnest(gs_out) |>
  dplyr::select(-c(alevel, blevel)) |>
  flextable::flextable() |>
  flextable::autofit() |>
  flextable::hline(i = c(5, 10))

```
</div>

<div style="float: left; width: 48%;">
- Summary from `gsDesign2::gs_power_ahr` for $\alpha = 0.0125$:

```{r}
os_gs2_a125 |>
  summary() |>
  as_gt()
```
</div>

<div style="float: right; width: 48%;">
- Summary from `gsDesign2::gs_power_ahr` for $\alpha = 0.025$:

```{r}
os_gs2_a25 |>
  summary() |>
  as_gt()
```
</div>

### Summary of comparison between EAST, `gsDesign` and `gsDesign2`

The differences in the results between EAST, `gsDesign` and `gsDesign2` are presented below up to 4th decimal place. The values highlighted in green correspond to non-zero differences.

- Comparison between EAST vs `gsDesign` and `gsDesign2` for $\alpha = 0.0125$:


```{r, echo=FALSE}
#helper function to align the gsDesign2 summary with gsDesign summary
as_gs <- function(data){
  
  data <- data$bounds
  
  data |>
    dplyr::filter(Bound == 'Upper') |>
    tidyr::pivot_longer(cols = -c(Analysis, Bound), names_to = 'Value', values_to = 'Efficacy') |>
    dplyr::mutate(Value = dplyr::case_when(
      Value == '~HR at bound' ~ 'HR at bound',
      Value == 'Nominal p' ~ 'p (1-sided)',
      Value == 'Probability' ~ 'P(Cross) if HR=0.60',
      Value == 'Probability0' ~ 'P(Cross) if HR=1',
      TRUE ~ Value
    ),
    Value_index = dplyr::case_when(
      Value == 'HR at bound' ~ 3,
      Value == 'p (1-sided)' ~ 2,
      Value == 'P(Cross) if HR=0.60' ~ 5,
      Value == 'P(Cross) if HR=1' ~ 4,
      TRUE ~ 1
    )
    ) |>
    dplyr::arrange(Analysis, Value_index) |>
    dplyr::select(Analysis, Value, Efficacy)
  
}
```


```{r}
os_gs |>
  dplyr::filter(alevel == 0.0125) |>
  tidyr::unnest(gs_out) |>
  dplyr::select(-c(alevel, blevel)) |>
  dplyr::select(eff_gs = Efficacy) |>
  dplyr::bind_cols(os_east |>
                     dplyr::select(value, analysis, eff_east = eff_125) |>
                     dplyr::mutate(eff_east = ifelse(value == 'Z', -eff_east, eff_east))) |>
  dplyr::bind_cols(as_gs(os_gs2_a125) |>
                     dplyr::select(eff_gs2 = Efficacy)) |>
  dplyr::mutate(
    diff_gsDesign = abs(eff_gs - eff_east),
    diff_gsDesign2 = abs(eff_gs2 - eff_east)
    ) |>
  dplyr::select(analysis, value, diff_gsDesign, diff_gsDesign2) |>
  flextable::flextable() |>
  flextable::autofit() |>
    flextable::set_header_labels(
    analysis = 'Analysis',
    value = 'Value',
    diff_gsDesign = 'EAST vs gsDesign ',
    diff_gsDesign2 = 'EAST vs gsDesign2'
  ) |>
  flextable::hline(i = c(5, 10)) |>
  flextable::colformat_double(j = c(3, 4), digits = 4) |>
  flextable::bg(~ round(diff_gsDesign, 4) != 0,  ~ diff_gsDesign, bg = '#6ECEB2') |>
  flextable::bg(~ round(diff_gsDesign2, 4) != 0,  ~ diff_gsDesign2, bg = '#6ECEB2')
```

- Comparison between EAST vs `gsDesign` and `gsDesign2` for $\alpha = 0.025$:

```{r}
os_gs |>
  dplyr::filter(alevel == 0.025) |>
  tidyr::unnest(gs_out) |>
  dplyr::select(-c(alevel, blevel)) |>
  dplyr::select(eff_gs = Efficacy) |>
  dplyr::bind_cols(os_east |>
                     dplyr::select(value, analysis, eff_east = eff_25) |>
                     dplyr::mutate(eff_east = ifelse(value == 'Z', -eff_east, eff_east))) |>
  dplyr::bind_cols(as_gs(os_gs2_a25) |>
                     dplyr::select(eff_gs2 = Efficacy)) |>
  dplyr::mutate(
    diff_gsDesign = abs(eff_gs - eff_east),
    diff_gsDesign2 = abs(eff_gs2 - eff_east)
    ) |>
  dplyr::select(analysis, value, diff_gsDesign, diff_gsDesign2) |>
  flextable::flextable() |>
  flextable::autofit() |>
    flextable::set_header_labels(
    analysis = 'Analysis',
    value = 'Value',
    diff_gsDesign = 'EAST vs gsDesign ',
    diff_gsDesign2 = 'EAST vs gsDesign2'
  ) |>
  flextable::hline(i = c(5, 10)) |>
  flextable::colformat_double(j = c(3, 4), digits = 4) |>
  flextable::bg(~ round(diff_gsDesign, 4) != 0,  ~ diff_gsDesign, bg = '#6ECEB2') |>
  flextable::bg(~ round(diff_gsDesign2, 4) != 0,  ~ diff_gsDesign2, bg = '#6ECEB2')
```

# Interim Analysis for Efficacy and Futility

