---
title: "gs_info_ahr: compute statistical information by the AHR method"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gs_info_ahr: compute statistical information by the AHR method}
---


```{r, essage=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.align = 'center') 
```

```{r, message=FALSE, echo=FALSE}
library(gt)
library(tibble)
library(dplyr)
library(testthat)
devtools::load_all()
#library(gsDesign2)
```

# Introduction of `gs_info_ahr()`

`tEvents()` calculate the analysis time (`Time` in its output), number of events (`Events` in its output), average hazard ratio (`AHR` in its outputs), effect size (`theta` in its output), statistical information (`info` and `info0` in its output) using an average hazard ratio model.

The aforementioned calculation is based on piecewise model:
+ piecewise constant enrollment rates 
+ piecewise exponential failure rates
+ piecewise censoring rates.


# Use Cases

## Example 1 

In this example, we only input the target number of events by `events = ...`, and derive the time when these events will be arrived.
```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1

gs_info_ahr(enrollRates = enrollRates, failRates = failRates,
            ratio = ratio, events = c(50, 80, 100))
```

## Example 2

In this example, we only input the analysis time by `analysisTimes = ...`, and derive the number of events at these analysis time.
```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1

gs_info_ahr(enrollRates = enrollRates, failRates = failRates,
            ratio = ratio, analysisTimes = c(10, 15, 20))
```

## Example 3

In this example, we both input `analysisTimes = ...` and `events = ...`. 
In this case, one will see 
+ the derived analysis time (`Time` column) $\geq$ input `analysisTimes`
+ the derived number of event (`Events` column) $\geq$ input `events`

```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1

gs_info_ahr(enrollRates = enrollRates, failRates = failRates,
            ratio = ratio, analysisTimes = c(10, 15, 20), events = c(80,  # > events in example 2 
                                                                     140, # < > events in example 2 
                                                                     220  # > events in example 2 
                                                                     ))
```





# Inner Logic of `gs_info_ahr()`

To explain the inner logic of `gs_info_ahr()`, we discuss 3 scenario.

1. only input `analysisTimes`
1. only input `events`
1. both input `analysisTimes` and `events`

## Scenario 1: only input `analysisTimes`

If only `analysisTimes = ...` is input, essentially, `gs_info_ahr()` uses `AHR()` to calculate the number of events at these `analysisTimes`.
```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1
analysisTimes <- c(10, 15, 20)

AHR(enrollRates = enrollRates, failRates = failRates, 
    ratio = ratio, totalDuration = analysisTimes) %>% 
  mutate(theta = -log(AHR), Analysis = 1 : length(analysisTimes)) %>% 
  select(Analysis, Time, Events, AHR, theta, info, info0) %>% 
  gt()
```

This is exactly the output from `gs_info_ahr()`:
```{r}
gs_info_ahr(enrollRates = enrollRates, failRates = failRates, 
            ratio = ratio, analysisTimes = analysisTimes) %>% gt()
```


## Scenario 2: only input `events`

If only `events = ...` is input, essentially, `gs_info_ahr()` uses `tEvents()` to calculate the time when these events will be arrived.
```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1
events <- c(70, 150, 200)

ans <- NULL
for(i in seq_along(events)){
  ans_new <- gsDesign2::tEvents(enrollRates = enrollRates, failRates = failRates, 
                                ratio = ratio, targetEvents = events[i])
  ans <- rbind(ans, ans_new)
}

ans %>% 
  mutate(theta = -log(AHR), Analysis = 1 : length(analysisTimes)) %>% 
  select(Analysis, Time, Events, AHR, theta, info, info0) %>% 
  gt()
```

This is exactly the output from `gs_info_ahr()`:
```{r}
gs_info_ahr(enrollRates = enrollRates, failRates = failRates, 
            ratio = ratio, events = events) %>% gt()
```


## Scenario 3: both input `analysisTimes` and `events`

If both `analysisTimes = ...` and `events = ...` are input, `gs_info_ahr()` uses both `AHR()` and `tEvents()`. 
In this way, it is guaranteed that
+ the derived number of event (`Events` column) $\geq$ input `events`
+ the derived analysis time (`Time` column) $\geq$ input `analysisTimes`

```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1
analysisTimes <- c(10, 15, 20)
events <- c(70, 150, 200)

ans <- NULL

# first, use `AHR()` to calculate the number of events at the input `analysisTimes`
ans <- AHR(enrollRates = enrollRates, failRates = failRates, 
           ratio = ratio, totalDuration = analysisTimes)

# second, compare if the events derived above meet the targeted number of events input in `events`
for(i in seq_along(events)){
  if (ans$Events[i] < events[i]){
    ans[i,] <- tEvents(enrollRates = enrollRates, failRates = failRates, 
                       ratio = ratio, targetEvents = events[i])
  }
}

ans %>% 
  mutate(theta = -log(AHR), Analysis = 1 : length(analysisTimes)) %>% 
  select(Analysis, Time, Events, AHR, theta, info, info0) %>% 
  gt()
```

This is exactly the output from `gs_info_ahr()`:
```{r}
gs_info_ahr(enrollRates = enrollRates, failRates = failRates, 
            ratio = ratio, events = events, analysisTimes = analysisTimes) %>% gt()
```