---
title: "gs_b: specify fixed boundaries in group sequential designs"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gs_b: specify fixed boundaries in group sequential designs}
---

```{r, message=FALSE}
library(tibble)
library(dplyr)
#library(gsDesign2)
devtools::load_all()
```

# Introduction of `gs_b()`

`gs_b()` can be used to derive fixed boundary in fixed/group sequential design. It is usually used in the `upper = ...` and `lower = ...` arguments in 
+ `gs_power_npe()`
+ `gs_design_npe()`
+ `gs_power_ahr()`
+ `gs_design_ahr()`
+ `gs_power_wlr()`
+ `gs_design_wlr()`
+ `gs_power_combo()`
+ `gs_design_combo()`

# Usage of `gs_b()`

## Example 1
Assume it is a group sequential design with 3 analysis, one can input its upper bound as a vector `c(4, 3, 2)` by using `gs_b()` as follows.
```{r}
gs_b(par = 4:2)
```


## Example 2
In the above example, one can assign the upper bound at the second analysis by
```{r}
gs_b(par = 4:2, k = 2)
```

## Example 3
Generate an efficacy bound using a spending function. 
Use Lan-DeMets spending approximation of O'Brien-Fleming bound as 50\%, 75\% and 100\% of final spending
```{r}
# information fraction
IF <- c(.5, .75, 1)
# Lan-DeMets spending approximation of O'Brien-Fleming
par <- gsDesign::gsDesign(alpha = .025, k = length(IF), 
                          test.type = 1, sfu = gsDesign::sfLDOF, 
                          timing = IF)$upper$bound
gs_b(par = par)
```


# Inner Logic of `gs_b()`

`gs_b` is a short function with 2 key arguments: `par = ...` and `k = ...`
```{r, eval=FALSE}
if(is.null(k)){
  return(par)
}else{ 
  return(par[k])
}
```

