---
title: "AHR: computes AHR under NPH assumptions and (stratified) populations"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{AHR: computes AHR under NPH assumptions and (stratified) populations}
---

```{r, message=FALSE}
library(tibble)
library(dplyr)
#library(gsDesign2)
devtools::load_all()
```

# Introduction of `AHR()`

`AHR()` provides a **geometric average hazard ratio** under various non-proportional hazards assumptions for either single or multiple strata studies. 
The piecewise exponential distribution allows a simple method to specify a distribution and enrollment pattern where the enrollment, failure and dropout rates changes over time.

# Usage of `AHR()`

## Example 1: Un-stratified population
```{r}
enrollRates <- tibble(Stratum = "All",
                      duration = c(2, 10, 4, 4, 8),
                      rate = c(5, 10, 0, 3, 6))
failRates <- tibble(Stratum = "All",
                    duration = 1,
                    failRate = c(.1, .2, .3, .4),
                    hr = c(.9, .75, .8, .6),
                    dropoutRate = .001)
AHR(enrollRates = enrollRates, failRates = failRates, totalDuration = c(15, 30))
```


## Example 2: Stratified population
```{r}
enrollRates <- tibble(Stratum = c(rep("Low", 2), rep("High", 3)),
                      duration = c(2, 10, 4, 4, 8),
                      rate = c(5, 10, 0, 3, 6))
failRates <- tibble(Stratum = c(rep("Low", 2), rep("High", 2)),
                    duration = 1,
                    failRate = c(.1, .2, .3, .4),
                    hr = c(.9, .75, .8, .6),
                    dropoutRate = .001)
AHR(enrollRates = enrollRates, failRates = failRates, totalDuration = c(15, 30))
```

# Inner Logic of `AHR()`

Let's take the un-stratified population as an example, where the enrollment rate, failure rates and dropout rates are
```{r}
enrollRates <- tibble(Stratum = "All",
                      duration = c(2, 10, 4),
                      rate = c(5, 10, 0))
failRates <- tibble(Stratum = "All",
                    duration = 1,
                    failRate = c(.1, .2),
                    hr = c(.9, .75),
                    dropoutRate = .001)

ratio <- 2

totalDuration <- 30
```

**Step 1:** compute proportion in each group
```{r}
Qe <- ratio / (1 + ratio)
Qc <- 1 - Qe
```

```{r, echo=FALSE}
cat("The proportion of the experimental arm is ", Qe, "\n")
cat("The proportion of the control arm is ", Qc, "\n")
```
To compute the expected events over different treatment group, stratum and time period, we iterate over `totalDuration` and `Strata`.
Since in this example, we only have one analysis time (`totalDuration = 30`) and one stratum (`Stratum = "All"`), we only iterate once. 
In one has multiple analysis time and strata, one can use a for loop and bind the results by row. 

```{r}
td <- totalDuration
s <- "All"
```



**Step 2:** subset the enrollment rates and failure rates by stratum.
```{r}
enroll <- enrollRates %>% filter(Stratum == s)
fail <- failRates %>% filter(Stratum == s)
```

**Step 3:** we calculate the enrollment rates in experimental arm and control arm, respectively.
```{r}
enroll_c <- enroll %>% mutate(rate = rate * Qc)
enroll_e <- enroll %>% mutate(rate = rate * Qe)
```

**Step 4:** we update the failure rate in the control and experimental arm.
```{r}
fail_c <- fail
fail_e <- fail %>% mutate(failRate = failRate * hr)
```

**Step 5:** we calculate the expected number of events in the control and experimental by `eEvents_df()`.
```{r}
events_c <- eEvents_df(enrollRates = enroll_c, failRates = fail_c, totalDuration = td, simple = FALSE)
events_e <- eEvents_df(enrollRates = enroll_e, failRates = fail_e, totalDuration = td, simple = FALSE)
```

```{r, echo=FALSE}
cat("The expected number of events in the control arm is \n")
events_c

cat("The expected number of events in the experimental arm is \n")
events_e
```


Here the `t` column is the start of period, the `failRate` column is the failure rate during the period, and the `Events` column is the  expected events during the period.

**Step 6:** we combine the results together and output it.
```{r, message=FALSE}
          # combine control and experimental
events <- rbind(events_c %>% mutate(Treatment = "Control"),
                events_e %>% mutate(Treatment = "Experimental")) %>%
          arrange(t, Treatment) %>% 
          ungroup() %>% 
          # recompute HR, events, info by period
          group_by(t) %>%
          summarize(Stratum = s, 
                    info = (sum(1 / Events))^(-1),
                    Events = sum(Events), 
                    HR = last(failRate) / first(failRate)) %>% 
          # compute info0
          mutate(Time = td, 
                 lnhr = log(HR), 
                 info0 = Events * Qc * Qe) %>%
          ungroup() %>% 
          
          group_by(Time, Stratum, HR) %>%
          summarize(t = min(t), 
                    Events = sum(Events), 
                    info0 = sum(info0), 
                    info = sum(info)) %>%
          # pool time period together
          group_by(Time) %>%
          summarize(AHR = exp(sum(log(HR) * Events) / sum(Events)),
                    Events = sum(Events),
                    info = sum(info),
                    info0 = sum(info0))
```

```{r, echo=FALSE}
cat("The overall expected number of events over the time is \n")
events
```

Please note that, in the output, the `info` column is based on the following input.
```{r, eval=FALSE}
enrollRates <- tibble(Stratum = "All",
                      duration = c(2, 10, 4),
                      rate = c(5, 10, 0))
failRates <- tibble(Stratum = "All",
                    duration = 1,
                    failRate = c(.1, .2),
                    hr = c(.9, .75),
                    dropoutRate = .001)
```
If the alternative hypothesis $H_1$ is 
$$
  \text{hr}
  =
  \left\{
  \begin{array}{ll}
    0.9  & \text{for the first 1 month} \\
    0.75 & \text{afterwards},
  \end{array}
  \right.
$$
then `info = info1`, where `info1` is the statistical information under $H_1$. 
But notice that the above `enrollRates` and `failRates` is not always the $H_1$, so we call it as `info`, rather than `info1`. 