---
title: "gs_power_npe: derives bounds and crossing probabilities for group sequential designs under NPH assumptions"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gs_power_npe: derives bounds and crossing probabilities for group sequential designs under NPH assumptions}
---

```{r, message=FALSE}
library(tibble)
library(dplyr)
library(gt)
#library(gsDesign2)
devtools::load_all()
```

# Introduction of `gs_power_npe()`

`gs_power_npe()` derives group sequential **bounds** and boundary crossing **probabilities** for a design. 
It allows a **non-constant treatment effect** over time, but also can be applied for the usual homogeneous effect size designs. 
It requires 
+ treatment effect (`theta`, `theta1`)
+ statistical information at each analysis (`info`, `info0`, `info1`)
+ a method of deriving bounds, such as fixed bounds or spending (`upper`, `upar`, `lower`, `lpar`). 

The routine enables two things not available in the gsDesign package:

1. non-constant effect, 
1. more flexibility in boundary selection. 



# Usage of `gs_power_npe()`

## Example 1: Fixed bound {.tabset}

### no futility bound

```{r}
# Same fixed efficacy bounds,  (i.e., non-binding bound), null hypothesis
gs_power_npe(
  theta = rep(0, 3),
  info = (1:3) * 40,
  upar = gsDesign::gsDesign(k = 3,sfu = gsDesign::sfLDOF)$upper$bound,
  lpar = rep(-Inf, 3)) %>%
  filter(Bound == "Upper") %>% gt()
```

### with futility bound 
```{r}
# Fixed bound
gs_power_npe(
  theta = c(.1, .2, .3),
  info = (1:3) * 40,
  upper = gs_b,
  upar = gsDesign::gsDesign(k = 3,sfu = gsDesign::sfLDOF)$upper$bound,
  lower = gs_b,
  lpar = c(-1, 0, 0)) %>% gt()
```


### futility only at analysis 1
```{r}
gs_power_npe(
  theta = c(.1, .2, .3),
  info = (1:3) * 40,
  upper = gs_b,
  upar = c(Inf, 3, 2),
  lower = gs_b,
  lpar = c(qnorm(.1), -Inf, -Inf)) %>% gt()
```


## Example 2: spending bounds {.tabset}

### lower spending based on non-zero effect
```{r}
gs_power_npe(
  theta = c(.1, .2, .3), # non-zero effect
  info = (1:3) * 40,
  upper = gs_spending_bound,
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  lower = gs_spending_bound,
  lpar = list(sf = gsDesign::sfHSD, total_spend = 0.1, param = -1, timing = NULL))
```

### 2-sided symmetric spend
```{r}
x <- gs_power_npe(
  theta = rep(0, 3),
  info = (1:3) * 40,
  # typically, 2-sided bounds are binding
  binding = TRUE,
  upper = gs_spending_bound,
  # O'Brien-Fleming spending
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  lower = gs_spending_bound,
  lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL))

x %>% gt()
```

```{r}
# Re-use these bounds under alternate hypothesis
# Always use binding = TRUE for power calculations
gs_power_npe(
  theta = c(.1, .2, .3),
  info = (1:3) * 40,
  binding = TRUE,
  upar = (x %>% filter(Bound == "Upper"))$Z,
  lpar = -(x %>% filter(Bound == "Upper"))$Z) %>% gt()
```


# Inner Logic of `gs_spending_bound()`

TODO


