---
title: "fixed_design: compute sample size/power of a fixed design"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    highlight: "textmate"
    css: "custom.css"
bibliography: ggsd.bib
vignette: |
  %\VignetteIndexEntry{fixed_design: compute sample size/power of a fixed design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r, message=FALSE, warning=FALSE}
library(tibble)
library(gt)
library(gsDesign)
library(gsDesign2)
```


# Introduction

In this vignette, we introduce the key functionality of `fixed_design` in `gsDesign2`. 
Generally speaking, `fixed_design` is caplable to both power calculations and sample size calculations under many methods, including

- AHR
- Fleming-Harrington (FH) [@FarringtonManning]
- Magirr-Burman (MB)
- Lachin-Foulkes (LF) [@LachinFoulkes]
- RMST [@YungLiu]
- Milestone [@YungLiu]
- risk difference (RD) [@MehrotraRailkar].

Throughout this vignette, we assume the design parameter is 
```{r}
# Enrollment rate
enrollRates <- tibble::tibble(
   Stratum = "All", 
   duration = 18, 
   rate = 20)

# Failure rates
failRates <- tibble::tibble(
   Stratum = "All", 
   duration = c(4, 100), 
   failRate = log(2) / 12,
   hr = c(1, .6), 
   dropoutRate = .001)

# Study duration in months
studyDuration <- 36

# Experimental / Control randomization ratio
ratio <- 1 

# 1-sided Type I error
alpha <- 0.025 
# Type II error (1 - power)
beta <- 0.1 
```

# Examples of AHR {.tabset}

If one wants to get a power of 90%, one can calculate the sample size by
```{r}
fixed_design(x = "AHR", 
             alpha = alpha, power = 1 - beta, 
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio) %>% 
  summary() %>% 
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of AHR by
```{r}
fixed_design(x = "AHR", 
             alpha = alpha,  
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio) %>% 
  summary() %>% 
  as_gt()
```

# Examples of FH {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
# fixed design with a given power with default rho/gamma
fixed_design(x = "FH", 
             alpha = alpha, power = 1 - beta, 
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio) %>% 
  summary() %>% 
  as_gt()
```
In the above example, the FH parameters (`rho`, `gamma`) take the default value as `rho = 0`, `gamma = 0.5`. But users can always custom their `rho` and `gamma` by  
```{r}
# fixed design with a given power with input rho/gamma
fixed_design(x = "FH", 
             alpha = alpha, power = 1 - beta, 
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio,
             rho = 0.5, gamma = 0.5) %>% 
  summary() %>%  
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of FH under the default `rho`, `gamma` by
```{r}
# fixed design with power calculated
fixed_design(x = "FH", 
             alpha = alpha,  
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio) %>% 
  summary() %>%  
  as_gt()
```
But users can always custom their `rho` and `gamma` by  
```{r}
# fixed design with power calculated
fixed_design(x = "FH", 
             alpha = alpha,  
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio,
             rho = 0.5, gamma = 0.5) %>% 
  summary() %>%  
  as_gt()
```

# Examples of MB {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
fixed_design(x = "MB", 
             ratio = ratio, 
             alpha = alpha, power = 1 - beta,
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration) %>% 
  summary() %>% 
  as_gt()
```
In the above example, the MB parameter (`tau`) takes the default value as `tau = 6`. But users can always custom their `tau` by  
```{r}
fixed_design(x = "MB", 
             ratio = ratio, 
             alpha = alpha, power = 1 - beta,
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration,
             tau = 4) %>% 
  summary() %>%  
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of MB under the default `tau = 6` by
```{r}
fixed_design(x = "MB",
             ratio = ratio,
             alpha = alpha,
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration) %>% 
  summary() %>%  
  as_gt()
```
But users can always custom their `tau` by  
```{r}
fixed_design(x = "MB",
             ratio = ratio,
             alpha = alpha,
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration,
             tau = 4) %>% 
  summary() %>%  
  as_gt()
```

# Examples of LF {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
fixed_design(x = "LF", alpha = alpha, power = 1 - beta, 
             ratio = ratio,
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration) %>% 
  summary() %>%  
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of LF by
```{r}
fixed_design(x = "LF", alpha = alpha,  
             ratio = ratio, 
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration) %>% 
  summary() %>%  
  as_gt()
```

# Examples of MaxCombo {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
fixed_design(x = "MaxCombo", alpha = alpha, power = 1 - beta,
             ratio = ratio, 
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration) %>% summary() %>% as_gt()
```
In the above example, the MaxCombo parameters take the default value 
```{r}
data.frame(rho = c(0, 0), gamma = c(0, 0.5), tau = -1) %>% gt()
```
But users can always custom their `rho`, `gamma`, `tau` by  
```{r}
fixed_design(x = "MaxCombo", alpha = alpha, power = 1 - beta,
             ratio = ratio, 
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration,
             rho = c(0, 0.5, 0.5),
             gamma = c(0, 0, 0.5),
             tau = c(-1, 4, 6)) %>% 
  summary() %>% 
  as_gt()
```


Given the above enrollment rate and enrollment duration, one can calculate the power of MaxCombo under the default parameters by
```{r}
fixed_design(x = "MaxCombo", alpha = alpha, 
             ratio = ratio, 
             enrollRates = enrollRates,
             failRates = failRates,
             studyDuration = studyDuration) %>% 
  summary() %>% 
  as_gt()
```
But users can always custom their `rho`, `gamma`, `tau` by 
```{r}
fixed_design(x = "MaxCombo", alpha = alpha, 
             ratio = ratio, 
             enrollRates = enrollRates,
             failRates = failRates, 
             studyDuration = studyDuration,
             rho = c(0, 0.5, 0.5),
             gamma = c(0, 0, 0.5),
             tau = c(-1, 4, 6)) %>% 
  summary() %>% 
  as_gt()
```



# Examples of RMST {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
fixed_design(x = "RMST", 
             alpha = alpha, power = 1 - beta, 
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio,
             tau = 18) %>% summary() %>% as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of RMST by
```{r}
fixed_design(x = "RMST", 
             alpha = alpha,  
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio,
             tau = 18) %>% 
  summary() %>% 
  as_gt()
```


# Examples of Milestone {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
fixed_design(x = "Milestone", 
             alpha = alpha, power = 1 - beta, 
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio,
             tau = 18) %>% 
  summary() %>% 
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of Milestone by
```{r}
fixed_design(x = "Milestone", 
             alpha = alpha,  
             enrollRates = enrollRates, failRates = failRates, 
             studyDuration = studyDuration, ratio = ratio,
             tau = 18) %>% 
  summary() %>% 
  as_gt()
```


# Examples of RD {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by
```{r}
fixed_design(x = "RD", 
             alpha = alpha, power = 1 - beta, 
             p_c = .15, p_e = .1, rd0 = 0,
             ratio = ratio) %>% 
  summary() %>% 
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power by
```{r}
fixed_design(x = "RD", 
             alpha = alpha, power = NULL, 
             p_c = .15, p_e = .1, rd0 = 0,
             N = 2000, ratio = ratio) %>% 
  summary() %>% 
  as_gt()
```

# Examples to get multiple designs together

```{r, message = FALSE}
x_AHR <- fixed_design(x = "AHR", alpha = alpha, ratio = ratio,
                      enrollRates = enrollRates, failRates = failRates, 
                      studyDuration = studyDuration)

x_FH <- fixed_design(x = "FH", alpha = alpha, ratio = ratio, 
                     enrollRates = enrollRates, failRates = failRates, studyDuration = studyDuration,
                     rho = 0.5, gamma = 0.5) 

x_MB <- fixed_design(x = "MB", alpha = alpha, ratio = ratio, 
                     enrollRates = enrollRates,failRates = failRates, studyDuration = studyDuration, 
                     tau = 4)

x_LF <- fixed_design(x = "LF", alpha = alpha, ratio = ratio, 
                     enrollRates = enrollRates, failRates = failRates, studyDuration = studyDuration)

x_MaxCombo <- fixed_design(x = "MaxCombo", alpha = alpha, ratio = ratio, 
                           enrollRates = enrollRates, failRates = failRates, studyDuration = studyDuration,
                           rho = c(0, 0.5, 0.5), gamma = c(0, 0, 0.5), tau = c(-1, 4, 6))

x_RMST <- fixed_design(x = "RMST", alpha = alpha, ratio = ratio, 
                       enrollRates = enrollRates, failRates = failRates, studyDuration = studyDuration,
                       tau = 30)

x_Milestone <- fixed_design(x = "Milestone", alpha = alpha, ratio = ratio, 
                       enrollRates = enrollRates, failRates = failRates, studyDuration = studyDuration,
                       tau = 30)

rbind(summary(x_AHR), summary(x_FH), summary(x_MB), summary(x_LF), summary(x_MaxCombo), summary(x_RMST), summary(x_Milestone)) %>% gt()
```

