---
title: "summary & as_gt: summarize group sequential design"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: hide
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
bibliography: gsDesign.bib
vignette: |
  %\VignetteIndexEntry{summary & as_gt: summarize group sequential design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tibble)
library(gt)
devtools::load_all()
```

# Overview

This vignette introduces publication quality table production for group sequential designs in the **gsDesign2** package.  
It also demonstrates designs for an example scenario using multiple design approaches.
We divide the document into 3 parts:

- Design specification and derivation
- Printing design summary tables
- Details of output from design functions
- Details on table output options

The reader can decide which of these sections is of interest to them.

The function used to generate bounds tables is `gsDesign2::summary()`. 
Users can use `gsDesign2::as_gt()` to format the above table using the **gt** package. 

In this vignette, we introduce a general approach to bound summaries by examples using different design approaches for a time-to-event outcome:

- the average hazard ratio (AHR) method extended from @Mukhopadhyay2020 using `gsDesign2::gs_design_ahr()`;
- the weighted logrank (WLR) method of @Yung2019Bcs using `gsDesign2::gs_design_wlr()`;

# Design Specification and Derivation

## Design Parameters

The design parameters we use across the different designs derived are: 

```{r}
# enrollment/failure rates
enrollRates <- tibble::tibble(
   Stratum = "All", 
   duration = 12, 
   rate = 1)
failRates <- tibble::tibble(
   Stratum = "All", 
   duration = c(4, 100), 
   failRate = log(2) / 12,
   hr = c(1, .6), 
   dropoutRate = .001)

# Information fraction
IF <- (1:3)/3 
# Analysis times in months; first 2 will be ignored as IF will not be achieved
analysisTimes <- c(.01, .02, 36)  

# Experimental / Control randomization ratio
ratio <- 1 

# 1-sided Type I error
alpha <- 0.025 
# Type II error (1 - power)
beta <- 0.1 

# Upper bound
upper <- gsDesign2::gs_spending_bound     # alpha-spending bound
upar <- list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL)
           
# Lower bound
lower <- gsDesign2::gs_spending_bound    # beta-spending bound
lpar <- list(sf = gsDesign::sfHSD, total_spend = 0.1, param = 0, timing = NULL)

# Fleming-Harrington (FH) weight functions for weighted logrank (WLR)
wgt00 <- function(x, arm0, arm1){    # Equal weighting for logrank
   gsDesign2:::wlr_weight_fh(x, arm0, arm1, rho = 0, gamma = 0)}
wgt05 <- function(x, arm0, arm1){    # Early downweighting with FH(0,.5) 
   gsDesign2:::wlr_weight_fh(x, arm0, arm1, rho = 0, gamma = .5)}

# Both of above tests for MaxCombo: logrank and FH(0,.5)
fh_test <- rbind(
   # Include logrank for all 3 analyses
   data.frame(rho = 0, gamma = 0, tau = -1, test = 1, Analysis = 1:3, analysisTimes = c(12, 24, 36)), 
   # Only include FH(0,.5) for analyses 2 and 3
   data.frame(rho = c(0, 0.5), gamma = 0.5, tau = -1, test = 2:3, Analysis = 3, analysisTimes = 36))
```

## Deriving Designs

### AHR Design Derivation 

Using the design parameters above, the AHR design is derived as follows:

By using the design parameters above, one can generate an AHR model by `gs_design_ahr` as

```{r, message=FALSE}
x_design_ahr <- gs_design_ahr(
   enrollRates = enrollRates,
   failRates = failRates,
   IF = IF, 
   analysisTimes = analysisTimes, 
   ratio = ratio, 
   alpha = alpha, 
   beta = beta, 
   upper = upper,
   upar = upar,
   lower = lower,
   lpar = lpar
)

x_power_ahr <- gs_power_ahr(
   enrollRates = x_design_ahr$enrollRates,
   failRates = x_design_ahr$failRates,
   events = c(100, 200, 400),
   analysisTimes = NULL,
   upper = upper,
   upar = upar,
   lower = lower,
   lpar = lpar
)
```


### WLR Design Derivation

```{r, message=FALSE}
x_design_wlr <- gs_design_wlr(
   enrollRates = enrollRates,
   failRates = failRates,
   weight = wgt05, 
   IF = NULL, 
   analysisTimes = sort(unique(x_design_ahr$analysis$Time)), 
   ratio = ratio, 
   alpha = alpha, 
   beta = beta,   
   upper = upper,
   upar = upar,
   lower = lower,
   lpar = lpar
)

# x_power_wlr <- gs_power_wlr(
#    enrollRates = x_design_wlr$enrollRates,
#    failRates = x_design_wlr$failRates,
#    weight = wgt05, 
#    events = c(100, 150, 250),
#    analysisTimes = NULL, 
#    upper = upper,
#    upar = upar,
#    lower = lower,
#    lpar = lpar
# )
```

# Default Summary Table Production

Instead of outputting 4 detailed tables (a table of enrollment rates, a table of failure rates, a table of analysis summary, a table of bounds summary), users can get a com pensive summary table by calling `summary(x)`, where `x` is the object returned either by `gs_design_ahr` or `gs_design_wlr`.
The `summary()` function produces an overall summary table for bounds for publication in a protocol. 

For example, the default output of `summary()` for the AHR method is
```{r}
x_design_ahr %>% 
  summary() %>% 
  gt::gt() %>% 
  gt::fmt_number(columns = c(3:6), decimals = 4)
```

Please note the `summary()` can also be applied to objected returned by `gs_power_ahr()`. 
For example,
```{r}
x_power_ahr %>% 
  summary() %>% 
  gt::gt() %>% 
  gt::fmt_number(columns = c(3:6), decimals = 4)
```


And the default output of `summary()` for the WLR method is

```{r}
x_design_wlr %>% 
  summary() %>% 
  gt::gt() %>% 
  gt::fmt_number(columns = c(3:6), decimals = 4)
```

Note that `summary()` can also be applied to summarize an object returned by `gs_power_wlr()`.
```{r, eval=FALSE}
summary(x_power_wlr) %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = c(3:6), decimals = 4)
```


# Detailed Summary Table Formatting

Here we demonstrate options for formatting analysis rows, bound rows as well as other table parameters such as titles, labels and footnotes.

## Custom the Variables to be Summaried for Each Analysis

In the above default table summary table generated by `summary(x)`, the variables used to summarize each analysis includes `Analysis`, `Time`, `N`(sample size), `Events`, `AHR`, and `IF` (information fraction). 
But users can customize these variables chosen using `analysis_vars = ...` and the corresponding decimals displayed using the argument `analysis_decimals = ...`. 
For example
```{r}
x_design_ahr %>% 
  summary(analysis_vars = c("N", "Events"),
          analysis_decimals = c(1, 1)) %>% 
  gt::gt() %>% 
  gt::fmt_number(columns = c(3:6), decimals = 4)
```

Please note that there is no need to input `"Analysis"` into `analysis_vars = ...` as it will always appear.

## Custom the Bound Names

Users can also customize the bound names. 
In the default output generated by `summary(x)`, the bound name is `c("Efficacy", "Futility")`, which can be changed into `c("A is better", "B is better")` for a 2-sided design by using the argument `bound_names = ...`.
For example,

```{r, message=FALSE}
x_design_ahr %>% 
  summary(bound_names = c("A is better", "B is better")) %>% 
  mutate_if(is.numeric, round, digits = 4) %>% 
  gt::gt() %>% 
  gt::fmt_number(columns = c(3:6), decimals = 4)
```

## Custom into a gt Table and Add Title/SubTitle/Footnotes/Spanners

Users can also use `as_gt()` to get the the above R table into a gt table.
Furthermore, they can edit the title/subtitle/spanner/footnotes of the gt table by using the arguments in `summary`. 

```{r}
summary(x_design_ahr) %>% 
   as_gt(title = "Summary of the Crossing Probability",
         subtitle = "by Using gs_design_ahr",
         colname_spanner = "Cumulative boundary crossing probability",
         colname_spannersub = c("Alternate hypothesis", "Null hypothesis"),
         footnote = list(content = c("approximate hazard ratio to cross bound.", 
                                     "gs_design_ahr is a function in gsDesign2.",
                                     "AHR is average hazard ratio; IF is information fraction."),
                         location = c("~HR at bound", NA, NA),
                         attr = c("colname", "subtitle", "analysis")))
```

The above objective can also be realized by using functions in the R package `gt` for custom design of table layout. 
We note that `as_gt()` always produces a `gt` object and, thus, can be further customized with **gt** package formatting functions.
In the future, we to support rich text format using a function `as_rtf()` in a fashion similar to `as_gt()`.

## Custom the Variables to Display 

Users can select the variables to be displayed in the summary table by using the argument `display_colunm = ...`.
```{r}
x_design_ahr %>% 
  summary() %>% 
  as_gt(display_columns = c("Analysis", "Bound", "Z", "Probability"))
```

## Custom Whether to Show Infinity Bound or Not 

Users have options to either show the infinity bounds or not by taking advantage of `display_inf_bound = ...`.
```{r}
x_design_ahr %>% 
  summary() %>% 
  as_gt(display_inf_bound = FALSE)
```



# Details of Output from Design/Power Functions

There are four components in the objects returned by either `gs_design_ahr()`/`gs_design_wlr()` or `gs_power_ahr()`/`gs_power_wlr()`: 
1. failure rates: a table summarizing failure rate and dropout rate.
1. enrollment rates: a table summarizing the enrollment rate.
1. bounds: a table summarize the bound of each analysis.
1. analysis: a table summarize the each analysis, with each one row for one analysis one hypothsis. 

## Failure Rates

The failure rates of different gsDesign object can be obtained by using `x$failRates`, where `x` is the object returned either by `gs_design_ahr` or `gs_design_wlr`. 
For example, the failure rates of the AHR design derivation can be returned by calling 
```{r}
x_design_ahr$failRates %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = 3:5, decimals = 4)
```

Please note that both `x_design_ahr` and `x_wlr` returns the same failure rates, which is the same as that inputted as `failRates`. 
To verify, let's take a look at the failure rate of the WLR design derivation, which are shown as below.

```{r}
x_design_wlr$failRates %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = 3:5, decimals = 4)
```


## Enrollment

The enrollment rate of a gs design derivation can be collected by using `x$failRates`, where `x` is the object returned either by `gs_design_ahr` or `gs_design_wlr`.
For example, the enrollment rates of the AHR/WLR design derivation is

```{r}
x_design_ahr$enrollRates %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = 3, decimals = 4)
```

```{r}
x_design_wlr$enrollRates %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = 3, decimals = 4)
```

It can be seen that, although the design derivation is different, the enrollment rate table share the same table structure, same enrollment period durations for each rate. 
Yet, the enrollment rates differ between designs only by a multiplicative constant.

## Analysis

The analysis summary table has the structure of one row per analysis per hypothesis. 
And columns can vary with different defaults for each design option. 
This type of tables are useful for understanding commonalities in how designs are summarized for different models.
To get analysis summary table, users can call `x$analysis`, where `x` is the object returned either by `gs_design_ahr` or `gs_design_wlr`.
For example, the analysis summary of the AHR/WLR design derivation is
```{r}
x_design_ahr$analysis %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = 2:8, decimals = 4)
```

```{r}
x_design_wlr$analysis %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = 2:8, decimals = 4)
```


## Bounds

The analysis summary table has the structure of One row per analysis per bound per hypothesis. Columns can vary with different defaults for each design option.
To get a bouns summary table, users can call `x$analysis`, where `x` is the object returned either by `gs_design_ahr` or `gs_design_wlr`.
For example, the bounds summary of the AHR/WLR design derivation is

```{r}
x_design_ahr$bounds %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = c(3, 5:7), decimals = 4) 
```

```{r}
x_design_wlr$bounds %>% 
   gt::gt() %>% 
   gt::fmt_number(columns = c(3, 5:7), decimals = 4)
```


# References
