---
title: "tEvents: compute time when a targeted number of events is made"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{tEvents: compute time when a targeted number of events is made}
---


```{r, essage=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.align = 'center') 
```

```{r, message=FALSE, echo=FALSE}
library(gt)
library(tibble)
library(dplyr)
library(testthat)
#library(gsDesign2)
devtools::load_all()
```

# Introduction of `tEvents`

`tEvents()` predicts time at which a targeted events is made. 
It is designed as a twins to `AHR()`: it matches input/output format with `AHR()`. 

# Use Cases

## Example 1: 
```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1

x <- tEvents(enrollRates = enrollRates, failRates = failRates,
             ratio = ratio, targetEvents = 200)

x %>% gt()
```

## Example 2: 

In this example, we verify `tEvents()` by `AHR()`.
```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1

x <- AHR(enrollRates = enrollRates, failRates = failRates, 
         ratio = ratio, totalDuration = 20)
cat("The number of events by 20 months is ", x$Events, ".\n")

y <- tEvents(enrollRates = enrollRates, failRates = failRates,
             ratio = ratio, targetEvents = x$Events)

cat("The time to get ", x$Events, " is ", y$Time, "months.\n")
```





# Inner Logic of `tEvents()`

The inner logic of `tEvents()` is to uniroot `AHR()` on `totalDuration`.

**Step 1:** find the difference between  `AHR()` and different values of `totalDuration`.
```{r}
foo <- function(x){
  ans <- AHR(enrollRates = enrollRates, failRates = failRates, 
             totalDuration = x, ratio = ratio)$Events - targetEvents
  return(ans)
}
```


```{r}
enrollRates <- tibble(Stratum = "All", duration = c(2, 2, 10), rate = c(3, 6, 9) * 5)
failRates <- tibble(Stratum = "All", duration = c(3, 100), failRate = log(2) / c(9, 18), hr = c(.9, .6), dropoutRate = rep(.001, 2))
ratio <- 1
targetEvents <- 200

cat("The difference between `targetEvents = 200` and the events after 30 months is ", foo(30), ".\n")
```


**Step 2:** uniroot `AHR()` on `totalDuration`.
```{r}
res <- uniroot(foo, interval = c(0.01, 100))

ans <- AHR(enrollRates = enrollRates, failRates = failRates, 
           totalDuration = res$root, ratio = ratio)
cat("After ", ans$Time, " months, there will be ", targetEvents, " events .\n")
```




