---
title: "Efficacy and futility boundary update"
author: "Yujie Zhao and Keaven M. Anderson"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
    code_folding: hide
bibliography: "gsDesign2.bib"
vignette: >
  %\VignetteIndexEntry{Efficacy and futility boundary update}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(gsDesign2)
library(gt)
```

# Design assumptions

We assume two analyses: an interim analysis (IA) and a final analysis (FA).
The IA is planned 20 months after opening enrollment, followed by the FA at
month 36.
The planned enrollment period spans 14 months, with the first 2 months having
an enrollment rate of 1/3 the final rate, the next 2 months with a rate of 2/3
of the final rate, and the final rate for the remaining 10 months.
To obtain the targeted 90\% power, these rates will be multiplied by a constant.
The control arm is assumed to follow an exponential distribution with a median
of 9 months and the dropout rate is 0.0001 per month regardless of treatment group.
Finally, the experimental treatment group is piecewise exponential with a
3-month delayed treatment effect; that is, in the first 3 months HR = 1 and
the HR is 0.6 thereafter.

```{r}
alpha <- 0.0125
beta <- 0.1
ratio <- 1

# Enrollment
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = (1:3) / 3
)

# Failure and dropout
fail_rate <- define_fail_rate(
  duration = c(3, Inf),
  fail_rate = log(2) / 9,
  hr = c(1, 0.6),
  dropout_rate = .0001
)
# IA and FA analysis time
analysis_time <- c(20, 36)

# Randomization ratio
ratio <- 1
```

We use the null hypothesis information for boundary crossing probability
calculations under both the null and alternate hypotheses.
This will also imply the null hypothesis information will be used for the
information fraction used in spending functions to derive the design.

```{r}
info_scale <- "h0_info"
```

# One-sided design 

For the design, we have efficacy bounds at both the IA and FA.
We use the @lan1983discrete spending function with a total alpha of `r alpha`,
which approximates an O'Brien-Fleming bound.

```{r}
upper <- gs_spending_bound
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)

x <- gs_design_ahr(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  alpha = alpha,
  beta = beta,
  info_frac = NULL,
  info_scale = "h0_info",
  analysis_time = analysis_time,
  ratio = ratio,
  upper = gs_spending_bound,
  upar = upar,
  test_upper = TRUE,
  lower = gs_b,
  lpar = rep(-Inf, 2),
  test_lower = FALSE
) |> to_integer()
```

The planned design targets:

- Planned events: `r round(x$analysis$event, 0)`
- Planned information fraction for interim and final analysis: `r round(x$analysis$info_frac, 4)`
- Planned alpha spending: `r round(gsDesign::sfLDOF(0.025, x$analysis$info_frac)$spend, 4)`
- Planned efficacy bounds: `r round(x$bound$z[x$bound$bound == "upper"], 4)`

We note that rounding up the final targeted events increases power slightly
over the targeted 90\%.

```{r}
x |>
  summary() |>
  as_gt() |>
  tab_header(title = "Planned design")
```

## Bounds for alternate alpha

At the stage of study design, we may be required to report the designs under multiple $\alpha$ if 
alpha is reallocated due to rejection of another hypothesis. At the design stage, the planned $\alpha$ is 0.0125. 
Assume the updated $\alpha$ is 0.025 due to reallocation of $\alpha$ from some other hypothesis. 
The corresponding bounds are
```{r}
gs_update_ahr(
  x = x,
  alpha = 0.025
  ) |>
  summary(col_decimals = c(z = 4)) |>
  as_gt(title = "Updated design",
        subtitle = "For alternate alpha = 0.025")
```
The above updated boundaries utilize the planned treatment effect and the planned statistical information under null hypothesis, considering the original design has `info_scale = "h0_info"`.

## Updating bounds with observed events at time of analyses

The planned events at interim analysis and final analysis are `r x$analysis$event`. 
However, at the time of each analysis, the observed events may differ from the planned events
due to using calendar-based cutoffs. For example, we assume that 240 and 352 events were observed at IA and FA, 
respectively. At the interim analysis, 30 events were observed during the delayed effect period, 
At the final analysis, 32 events were observed during the delayed effect period.
In practice, even if attempting to match event counts exactly the observed events at analyses often differ from planned.
We also assume the protocol specifies that the full $\alpha$ will be spent at
the final analysis even in a case like this when there is a shortfall of events
versus the design plan.


The updated design is
```{r}
# Set spending fraction for interim according to observed events 
# divided by planned final events.
# Final spending fraction is 1 per plan even if there is a shortfall
# of events versus planned (as specified above)
# Update bound
gs_update_ahr(
  x = x,
  ustime = c(240 / max(x$analysis$event), 1),
  event_tbl = data.frame(analysis = c(1, 1, 2, 2), 
                         event = c(30, 210, 32, 320))
) |>
  summary(col_decimals = c(z = 4)) |>
  as_gt(title = "Updated design",
        subtitle = paste0("With observed ", sum(observed_data_ia$event),
                          " events at IA and ", sum(observed_data_fa$event),
                          " events at FA"))
```

# Two-sided asymmetric design, beta-spending with non-binding lower bound

In this section, we investigate a 2 sided asymmetric design, with a
non-binding $\beta$-spending used to generate futility bounds. $\beta$-spending refers to
Type II error (1 - power) spending for the lower bound crossing probabilities under the
alternative hypothesis. Non-binding bound computation assumes the trial continues if the
lower bound is crossed for Type I error, but not Type II error.

In the original designs, we employ the Lan-DeMets spending function used to
approximate O'Brien-Fleming bounds [@lan1983discrete] for both efficacy and
futility bounds.
The total spending for efficacy is `r alpha`, and for futility is `r beta`.
In addition, we assume there is no futility test for the final analysis.

```{r}
# Upper and lower bounds uses spending with Lan-DeMets spending approximating
# O'Brien-Fleming bound
upper <- gs_spending_bound
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)
lower <- gs_spending_bound
lpar <- list(sf = gsDesign::sfLDOF, total_spend = beta, param = NULL)

x <- gs_design_ahr(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  alpha = alpha,
  beta = beta,
  info_frac = NULL,
  info_scale = "h0_info",
  analysis_time = c(20, 36),
  ratio = ratio,
  upper = gs_spending_bound,
  upar = upar,
  test_upper = TRUE,
  lower = lower,
  lpar = lpar,
  test_lower = c(TRUE, FALSE),
  binding = FALSE
) |> to_integer()
```

In the planned design, we have

- Planned events: `r round(x$analysis$event, 0)`
- Planned information fraction (timing): `r round(x$analysis$info_frac, 4)`
- Planned alpha spending: `r gsDesign::sfLDOF(0.025, x$analysis$info_frac)$spend`
- Planned efficacy bounds: `r round(x$bound$z[x$bound$bound == "upper"], 4)`
- Planned futility bounds: `r round(x$bound$z[x$bound$bound == "lower"], 4)`

Since we added futility bounds, the sample size and number of events are
larger than we had above in the 1-sided example.

```{r}
x |>
  summary() |>
  as_gt() |>
  tab_header(title = "Planned design",
             subtitle = "2-sided asymmetric design, non-binding futility")
```

## Bounds for alternate alpha

We may want to report the design bounds under multiple $\alpha$ 
in the case Type I error may be reallocated from another hypothesis. 
We assume now that $\alpha$ is 0.025 but we still use the same sample size and event timing as for the original alpha = 0.0125. The updated bounds are
```{r}
gs_update_ahr(
  x = x,
  alpha = 0.025
  ) |>
  summary(col_decimals = c(z = 4)) |>
  as_gt(title = "Updated design",
        subtitle = "For alpha = 0.025")
```

## Updating bounds with observed events at time of analyses

We assume the observed events same as for the 1-sided example above.

The updated design is
```{r}
# Update spending fraction as above
ustime <- c(sum(observed_data_ia$event) / max(x$analysis$event), 1)

gs_update_ahr(
  x = x,
  ustime = ustime,
  # Spending fraction for futility bound same as for efficacy
  lstime = ustime, 
  event_tbl = data.frame(analysis = c(1, 1, 2, 2), 
                         event = c(30, 210, 32, 320))
  ) |>
  summary(col_decimals = c(z = 4)) |>
  as_gt(title = "Updated design",
        subtitle = paste0("With observed 240 events at IA and 352 events at FA"))
```

# References
