---
title: "Futility bounds at design and analysis under non-proportional hazards"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
bibliography: "gsDesign2.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Futility bounds at design and analysis under non-proportional hazards}
---

```{r, warning=FALSE, message=FALSE}
library(gsDesign2)
library(gt)
library(dplyr)
library(tibble)
library(ggplot2)
```

# Overview

We set up futility bounds under a non-proportional hazards assumption.
We consider methods presented by @korn2018interim for setting such bounds and then consider an alternate futility bound based on $\beta-$spending under a delayed or crossing treatment effect to simplify implementation.
Finally, we show how to update this $\beta-$spending bound based on blinded interim data.
We will consider an example to reproduce a line of @korn2018interim Table 1 with the alternative futility bounds considered.

## Initial design set-up for fixed analysis

@korn2018interim considered delayed effect scenarios and proposed a futility bound that is a modification of an earlier method proposed by @wieand1994stopping.
We begin with the enrollment and failure rate assumptions which @korn2018interim based on an example by @chen2013statistical.

```{r}
# Enrollment assumed to be 680 patients over 12 months with no ramp-up
enroll_rate <- define_enroll_rate(duration = 12, rate = 680 / 12)
# Failure rates
## Control exponential with median of 12 mos
## Delayed effect with HR = 1 for 3 months and HR = .693 thereafter
## Censoring rate is 0
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = -log(.5) / 12,
  hr = c(1, .693),
  dropout_rate = 0
)
## Study duration was 34.8 in Korn & Freidlin Table 1
## We change to 34.86 here to obtain 512 expected events more precisely
study_duation <- 34.86
```

We now derive a fixed sample size based on these assumptions.
Ideally, we would allow a targeted event count and variable follow-up in `fixed_design()` so that the study duration will be computed automatically.

```{r}
fixedevents <- fixed_design(
  x = "ahr", alpha = 0.025, power = NULL,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duation = study_duation
)

fixedevents %>%
  summary() %>%
  select(-Bound) %>%
  as_gt(footnote = "Power based on 512 events") %>%
  fmt_number(columns = 3:4, decimals = 2) %>%
  fmt_number(columns = 5:6, decimals = 3)
```

# Modified Wieand futility bound

The @wieand1994stopping rule recommends stopping after 50% of planned events accrue if the observed HR > 1.
kornfreidlin2018 modified this by adding a second interim analysis after 75% of planned events and stop if the observed HR > 1
This is implemented here by requiring a trend in favor of control with a direction $Z$-bound at 0 resulting in the *Nominal p* bound being 0.5 for interim analyses in the table below.
A fixed bound is specified with the 'gs_b()` function for `upper` and `lower` and its correspoinding parameters `upar` for the upper (efficacy) bound and `lpar` for the lower (futility) bound.
The final efficacy bound is for a 1-sided nominal p-value of 0.025; the futility bound lowers this to 0.0247 as noted in the lower-right-hand corner of the table below.
In the last row under *Alternate hypothesis* below we see the power is 88.44%.
@korn2018interim computed 88.4% power for this design with 100,000 simulations which estimate the standard error for the power calculation to be `r paste(100 * round(sqrt(.884 * (1 - .884) / 100000), 4), "%", sep = "")`.

```{r}
wieand <- gs_power_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  upper = gs_b, upar = c(rep(Inf, 2), qnorm(.975)),
  lower = gs_b, lpar = c(0, 0, -Inf),
  event = 512 * c(.5, .75, 1)
)
wieand %>%
  summary() %>%
  as_gt(
    title = "Group sequential design with futility only at interim analyses",
    subtitle = "Wieand futility rule stops if HR > 1"
  )
```

# Beta-spending futility bound with AHR

Need to summarize here.

```{r}
betaspending <- gs_power_ahr(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  upper = gs_b,
  upar = c(rep(Inf, 2), qnorm(.975)),
  lower = gs_spending_bound,
  lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  event = 512 * c(.5, .75, 1),
  test_lower = c(TRUE, TRUE, FALSE)
)

betaspending %>%
  summary() %>%
  as_gt(
    title = "Group sequential design with futility only",
    subtitle = "Beta-spending futility bound"
  )
```

# Classical beta-spending futility bound

A classical $\beta-$spending bound would assume a constant treatment effect over time using the  proportional hazards assumption. We use the average hazard ratio at the fixed design analysis for this purpose.

# Korn and Freidlin futility bound

The @korn2018interim futility bound is set *when at least 50% of the expected events have occurred and at least two thirds of the observed events have occurred later than 3 months from randomization*.
The expected timing for this is demonstrated below.

## Accumulation of events by time interval

We consider the accumulation of events over time that occur during the no-effect interval for the first 3 months after randomization and events after this time interval.
This is done for the overall trial without dividing out by treatment group using the `gsDesign2::AHR()` function.
We consider monthly accumulation of events through the 34.86 months planned trial duration.
We note in the summary of early expected events below that all events during the first 3 months on-study are expected prior to the first interim analysis.

```{r}
event_accumulation <- pw_info(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  total_duration = c(1:34, 34.86),
  ratio = 1
)
head(event_accumulation, n = 7) %>% gt()
```

We can look at the proportion of events after the first 3 months as follows:

```{r}
event_accumulation %>%
  group_by(time) %>%
  summarize(`Total events` = sum(event), "Proportion early" = first(event) / `Total events`) %>%
  ggplot(aes(x = time, y = `Proportion early`)) +
  geom_line()
```

For the @korn2018interim bound the targeted timing is when both 50% of events have occurred and at least 2/3 are more than 3 months after enrollment with 3 months being the delayed effect period.
We see above that about 1/3 of events are still within 3 months of enrollment at month 20.

## Korn and Freidlin bound

The bound proposed by @korn2018interim

# References
