---
title: "Integer designs"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Integer designs}
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png"
)
options(width = 58)
```

```{r, message=FALSE}
library(gsDesign)
library(gsDesign2)
library(tibble)
library(dplyr)
library(gt)
```

# Unstratified design

## Binary outcome
```{r}
x <- gs_design_rd(
  p_c = tibble(stratum = "All", rate = .2),
  p_e = tibble(stratum = "All", rate = .15),
  info_frac = c(0.5, 0.8, 1),
  rd0 = 0,
  alpha = .025,
  beta = .1,
  ratio = 1,
  stratum_prev = NULL,
  weight = "unstratified",
  upper = gs_b,
  lower = gs_b,
  upar = gsDesign(k = 3, test.type = 1, sfu = sfLDOF)$upper$bound,
  lpar = c(qnorm(.1), rep(-Inf, 2))
)

x %>% summary() %>% as_gt()
```

```{r}
xi <- x %>% to_integer()

xi %>% summary() %>% as_gt()
```

Note that in the original design, the sample size is `r x$analysis$n`, and in the integer design, the sample size is updated to `r xi$analysis$n`. For the 2 interim analysis, we floor to the closet multiplier of 2, since the randomization ratio is `r x$input$ratio`. At the final analysis, we ceiling the sample size from `r dplyr::last(x$analysis$n)` to `r dplyr::last(xi$analysis$n)` and also make sure the integer sample size is a multiplier of 2.

Please also note that, since the sample size is rounded, the power of the new design also changes a little bit, i.e., from `r x$bound %>% filter(bound == "upper") %>% arrange(desc(analysis)) %>% filter(row_number() == 1) %>% select(probability) %>% unlist() %>% as.numeric()` to `r xi$bound %>% filter(bound == "upper") %>% arrange(desc(analysis)) %>% filter(row_number() == 1) %>% select(probability) %>% unlist() %>% as.numeric()`.

## Survival outcome
```{r}
x <- gs_design_ahr(analysis_time = c(12, 24, 36))
xi <- x %>% to_integer()
```

Notice that with the integer design, the (i) number of events, (ii) sample size, (iii) power, (iv) information fraction will be different.

```{r}
tibble(
  Design = rep(c("Original design", "Integer design"), each = 3),
  Events = c(x$analysis$event,  xi$analysis$event),
  `Sample size` = c(x$analysis$n, xi$analysis$n),
  `Information fraction` = c(x$analysis$info_frac, xi$analysis$info_frac),
  Power = c((x$bound %>% filter(bound == "upper"))$probability, (xi$bound %>% filter(bound == "upper"))$probability)
) %>%
  group_by(Design) %>%
  gt() %>%
  tab_header(title = "Comparision between the original/integer design",
             subtitle = "on survival endpoints") %>%
  fmt_number(columns = 2:5, decimals = 4)
```


# Stratified design

```{r}
y <- gs_design_rd(
  p_c = tibble(
    stratum = c("biomarker positive", "biomarker negative"),
    rate = c(.2, .25)
    ),
  p_e = tibble(
    stratum = c("biomarker positive", "biomarker negative"),
    rate = c(.15, .22)
    ),
  info_frac = c(0.7, 1),
  rd0 = 0,
  alpha = .025,
  beta = .1,
  ratio = 1,
  stratum_prev = tibble(
    stratum = c("biomarker positive", "biomarker negative"),
    prevalence = c(.4, .6)
    ),
  weight = "ss",
  upper = gs_spending_bound, lower = gs_b,
  upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
  lpar = rep(-Inf, 2)
  )

y %>% summary() %>% as_gt()
```

```{r}
yi <- y %>% to_integer()

yi %>% summary() %>% as_gt()
```

Note that in the original design, the sample size is `r y$analysis$n`, and in the integer design, the sample size is updated to `r yi$analysis$n`. For the 2 interim analysis, we floor to the closet multiplier of 2, since the randomization ratio is `r y$input$ratio`. At the final analysis, we ceiling the sample size from `r dplyr::last(y$analysis$n)` to `r dplyr::last(yi$analysis$n)` and also make sure the integer sample size is a multiplier of 2.
