---
title: "Trial design with spending under NPH"
author: "Keaven M. Anderson"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteIndexEntry{Trial design with spending under NPH}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tibble)
library(gsDesign)
library(gsDesign2)
library(gt)
```

# Overview

This vignette covers how to implement designs for trials with spending assuming non-proportional hazards.
We are primarily concerned with practical issues of implementation rather than design strategies, but we will not ignore design strategy.

# Scenario for Consideration

Here we set up enrollment, failure and dropout rates along with assumptions for enrollment duration and times of analyses.

In this example, we assume there are 4 analysis (3 interim analysis + 1 final analysis), and they are conducted after 18, 24, 30, 36 months after the trail starts.

```{r}
n_analysis <- 4
analysis_time <- c(18, 24, 30, 36)
```

And we further assume there is not stratum and the enrollment last for 12 months.
For the first 2 months, second 2 months, third 2 months and the reminder month, the enrollment rate is $8:12:16:24$.
Please note that $8:12:16:24$ is not the real enrollment rate.
Instead, it only specifies the enrollment rate ratio between different duration.

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 2, 6),
  rate = c(8, 12, 16, 24)
)

enroll_rate %>%
  gt() %>%
  tab_header(title = "Table of Enrollment")
```

Moreover, we assume the hazard ratio (HR) of the first 3 month is 0.9 and thereafter is 0.6.
We also assume the the survival time follow a piecewise exponential distribution with a median of 8 month for the first 3 months and 14 month thereafter.

```{r}
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(8, 14),
  hr = c(.9, .6),
  dropout_rate = .001
)

fail_rate %>%
  gt() %>%
  tab_header(title = "Table of Failure Rate")
```

# Fixed Design with no Interim Analysis

We can derive power for the above enrollment rates and failure rates as follows:

```{r}
fixed_design_ahr(enroll_rate = enroll_rate, fail_rate = fail_rate, power = NULL, ratio = 1, study_duration = 36, event = NULL) %>% summary()
```



We now compute sample size and then translate from a continuous sample size to an integer sample size.

```{r}
fixed_design <- fixed_design_ahr(enroll_rate = enroll_rate, fail_rate = fail_rate, power = .9, ratio = 1, study_duration = 36, event = NULL) 
integer_fixed_design <- fixed_design %>% to_integer()
# NOTE: the next line should not be needed, but to_integer() changed names
names(integer_fixed_design$analysis) <- names(fixed_design$analysis)
summary(integer_fixed_design)
```


## Group Sequential Design

We now consider a group sequential design with bounds derived using spending functions.
We target the interim analysis for 24 months and the final analysis for 36 months.
Spending for both efficacy and futility is based on the proportion of events expected at each analysis divided by the total expected events at the final analysis.



```{r}
gs <- 
  gs_design_ahr(
    enroll_rate = enroll_rate,
    fail_rate = fail_rate,
    info_frac = NULL,
    analysis_time = c(24, 36),
    upper = gs_spending_bound,
    lower = gs_spending_bound,
    upar = list(sf = gsDesign::sfLDOF, total_spend = 0.025, param = NULL, timing = NULL),
    lpar = list(sf = gsDesign::sfLDOF, total_spend = 0.1, param = NULL, timing = NULL), 
    h1_spending = TRUE
  ) %>% to_integer()
summary(gs) %>% gt()
```

