---
title: "Performance Optimization Tips for gsDesign2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Optimization Tips for gsDesign2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `gsDesign2` package provides powerful tools for group sequential design under non-proportional hazards. While the computational complexity of these designs can be substantial, there are several strategies to optimize performance.

## Key Performance Improvements

### Recent Performance Improvements

Recent versions include the following performance optimizations:

1. **Optimized Loop Operations**: Replaced inefficient `rbind` operations in loops with pre-allocated lists, reducing time complexity from O(n²) to O(n) in `gs_info_ahr()`.

2. **Improved Aggregation**: Replaced `lapply` + `split` + `do.call(rbind)` patterns with more efficient `aggregate()` operations in `expected_event()`.

3. **Function Memoization**: The package includes automatic caching of expensive function calls to avoid redundant computations.

## Performance Tips for Users

### 1. Use Vectorized Operations

When possible, specify multiple analysis times in a single call rather than calling functions repeatedly:

```{r, eval=FALSE}
# Efficient: Single call with multiple times
gs_info_ahr(analysis_time = c(18, 27, 36))

# Less efficient: Multiple individual calls
result1 <- gs_info_ahr(analysis_time = 18)
result2 <- gs_info_ahr(analysis_time = 27)
result3 <- gs_info_ahr(analysis_time = 36)
```

### 2. Enable Cache Reporting (Optional)

To monitor cache performance during development:

```{r, eval=FALSE}
options(gsDesign2.cache.report = TRUE)
# Your analysis code here
options(gsDesign2.cache.report = FALSE)
```

### 3. Reduce Complexity When Possible

- **Fewer strata**: Stratified designs are more computationally intensive. Use stratification only when necessary.
- **Fewer change points**: Piecewise constant rates with many change points increase computation time.
- **Calendar-based designs**: Using `analysis_time` is generally faster than event-driven designs with complex `event` and `analysis_time` combinations.

### 4. Parallel Processing for Multiple Scenarios

When evaluating multiple independent design scenarios, consider using parallel processing:

```{r, eval=FALSE}
library(parallel)

scenarios <- list(
  list(alpha = 0.025, beta = 0.1),
  list(alpha = 0.025, beta = 0.2),
  list(alpha = 0.01, beta = 0.1)
)

results <- mclapply(scenarios, function(s) {
  gs_design_ahr(alpha = s$alpha, beta = s$beta)
}, mc.cores = 3)
```

## Benchmarking Your Code

For critical analyses, benchmark your code to identify bottlenecks:

```{r, eval=FALSE}
library(tictoc)

tic("Design calculation")
design <- gs_design_ahr(
  analysis_time = c(12, 24, 36),
  alpha = 0.025,
  beta = 0.2
)
toc()
```

## Expected Performance

Typical performance for common operations (may vary by system):

- `gs_info_ahr()` with 3 analyses: 1-3 seconds
- `gs_design_ahr()` with 3 analyses: 3-10 seconds  
- `gs_design_combo()` with spending bounds: 10-30 seconds
- `gs_power_wlr()` with 3 analyses: 5-15 seconds

Complex designs with:
- Multiple strata
- Many piecewise rate changes
- Both event and time-driven analyses
- Spending function boundary derivation

...may require 30-60 seconds or more.

## Troubleshooting Slow Performance

If you experience unexpectedly slow performance:

1. **Check input complexity**: Simplify enrollment and failure rate specifications where possible.

2. **Verify convergence parameters**: The `interval` parameter in some functions affects root-finding performance. Tighten the interval if you know approximate solution bounds.

3. **Update to latest version**: Performance improvements are ongoing. Ensure you're using the latest package version:

```{r, eval=FALSE}
packageVersion("gsDesign2")
```

4. **Profile your code**: Use R's profiling tools to identify specific bottlenecks:

```{r, eval=FALSE}
Rprof("profile.out")
# Your code here
Rprof(NULL)
summaryRprof("profile.out")
```

## Reporting Performance Issues

If you encounter performance issues not addressed by these tips, please report them on GitHub with:

- A minimal reproducible example
- Your system specifications
- Package version information
- Expected vs. actual timing

This helps us continue improving the package for all users.

## Additional Resources

- See `inst/example_run_time/timer.Rmd` for detailed timing examples
- Check the package NEWS for recent performance improvements
- Visit our GitHub issues for known performance considerations
