---
title: "gs_info_ahr: compute statistical information by the AHR method"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gs_info_ahr: compute statistical information by the AHR method}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.align = "center")
```

```{r, message=FALSE, warning=FALSE}
library(gt)
library(dplyr)
library(testthat)
library(gsDesign2)
```

# Introduction of `gs_info_ahr()`

`expected_time()` calculate the analysis time (`Time` in its output), number of events (`Events` in its output), average hazard ratio (`AHR` in its outputs), effect size (`theta` in its output), statistical information (`info` and `info0` in its output) using an average hazard ratio model.

The aforementioned calculation is based on piecewise model:

- piecewise constant enrollment rates
- piecewise exponential failure rates
- piecewise censoring rates.

# Use Cases

## Example 1

In this example, we only input the target number of events by `event = ...`, and derive the time when these events will be arrived.
```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1

gs_info_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, event = c(50, 80, 100)
)
```

## Example 2

In this example, we only input the analysis time by `analysis_time = ...`, and derive the number of events at these analysis time.

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1

gs_info_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, analysis_time = c(10, 15, 20)
)
```

## Example 3

In this example, we both input `analysis_time = ...` and `event = ...`.
In this case, one will see

- the derived analysis time (`time` column) $\geq$ input `analysis_time`
- the derived number of event (`Events` column) $\geq$ input `event`

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1

gs_info_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, analysis_time = c(10, 15, 20), event = c(
    80, # > events in example 2
    140, # < > events in example 2
    220 # > events in example 2
  )
)
```

# Inner Logic of `gs_info_ahr()`

To explain the inner logic of `gs_info_ahr()`, we discuss 3 scenario.

1. only input `analysis_time`
1. only input `event`
1. both input `analysis_time` and `event`

## Scenario 1: only input `analysis_time`

If only `analysis_time = ...` is input, essentially, `gs_info_ahr()` uses `AHR()` to calculate the number of events at these `analysis_time`.

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1
analysis_time <- c(10, 15, 20)

ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, total_duration = analysis_time
) %>%
  mutate(theta = -log(ahr), Analysis = seq_along(analysis_time)) %>%
  select(Analysis, time, event, ahr, theta, info, info0) %>%
  gt()
```

This is exactly the output from `gs_info_ahr()`:

```{r}
gs_info_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, analysis_time = analysis_time
) %>% gt()
```

## Scenario 2: only input `event`

If only `event = ...` is input, essentially, `gs_info_ahr()` uses `expected_time()` to calculate the time when these events will be arrived.

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1
events <- c(70, 150, 200)

ans <- NULL
for (i in seq_along(events)) {
  ans_new <- gsDesign2::expected_time(
    enroll_rate = enroll_rate, fail_rate = fail_rate,
    ratio = ratio, target_event = events[i]
  )
  ans <- rbind(ans, ans_new)
}

ans %>%
  mutate(theta = -log(ahr), Analysis = seq_along(analysis_time)) %>%
  select(Analysis, time, event, ahr, theta, info, info0) %>%
  gt()
```

This is exactly the output from `gs_info_ahr()`:

```{r}
gs_info_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, event = events
) %>% gt()
```

## Scenario 3: both input `analysis_time` and `event`

If both `analysis_time = ...` and `event = ...` are input, `gs_info_ahr()` uses both `AHR()` and `expected_time()`.
In this way, it is guaranteed that

- the derived number of event (`event` column) $\geq$ input `event`
- the derived analysis time (`time` column) $\geq$ input `analysis_time`

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1
analysis_time <- c(10, 15, 20)
events <- c(70, 150, 200)

ans <- NULL

# First, use `AHR()` to calculate the number of events at the input `analysis_time`
ans <- ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, total_duration = analysis_time
)

# Second, compare if the events derived above meet the targeted number of events input in `events`
for (i in seq_along(events)) {
  if (ans$event[i] < events[i]) {
    ans[i, ] <- expected_time(
      enroll_rate = enroll_rate, fail_rate = fail_rate,
      ratio = ratio, target_event = events[i]
    )
  }
}

ans %>%
  mutate(theta = -log(ahr), Analysis = seq_along(analysis_time)) %>%
  select(Analysis, time, event, ahr, theta, info, info0) %>%
  gt()
```

This is exactly the output from `gs_info_ahr()`:

```{r}
gs_info_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, event = events, analysis_time = analysis_time
) %>% gt()
```
