---
title: "expected_accrual: computes the expected cumulative enrollment (accrual) given a set of piecewise constant enrollment rates and times"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{expected_accrual: computes the expected cumulative enrollment (accrual) given a set of piecewise constant enrollment rates and times}
---

```{r, message=FALSE, warning=FALSE}
library(tibble)
library(dplyr)
library(gsDesign2)
```

# Introduction of `expected_accrual()`


# Use cases of `expected_accrual()`

## Example 1
For the enrollment in the first 3 months, it is exactly $3 \times 5 = 15$.
```{r}
expected_accrual(
  time = 3,
  enroll_rate = define_enroll_rate(duration = c(3, 3, 18), rate = c(5, 10, 20))
)
```

## Example 2
For the enrollment in the first 6 months, it is exactly $3 \times 5 + 3 \times 10 = 45$.
```{r}
expected_accrual(
  time = 6,
  enroll_rate = define_enroll_rate(duration = c(3, 3, 18), rate = c(5, 10, 20))
)
```

## Example 3
For the enrollment in the first 24 months, it is exactly $3 \times 5 + 3 \times 10 + 18 * 20 = 405$.
```{r}
expected_accrual(
  time = 24,
  enroll_rate = define_enroll_rate(duration = c(3, 3, 18), rate = c(5, 10, 20))
)
```

## Example 4
For the enrollment after 24 months, it is the same as that from the 24 months, since the enrollment is stopped.
```{r}
expected_accrual(
  time = 25,
  enroll_rate = define_enroll_rate(duration = c(3, 3, 18), rate = c(5, 10, 20))
)
```

## Example 5
Instead of compute the enrolled subjects one time point by one time point, we can also compute it once.
```{r}
expected_accrual(
  time = c(3, 6, 24, 25),
  enroll_rate = define_enroll_rate(duration = c(3, 3, 18), rate = c(5, 10, 20))
)
```

# Inner Logic of `expected_accrual()`

TODO
