---
title: "Computing spending boundaries in group sequential design"
author: "Keaven Anderson and Yujie Zhao "
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
bibliography: "gsDesign2.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Computing spending boundaries in group sequential design}
---

```{r, message=FALSE, warning=FALSE}
library(tibble)
library(dplyr)
library(gsDesign2)
```

# Introduction

We compare derivation of different spending bounds using the *gsDesign2* and *gsDesign* packages.
In *gsDesign* there are 6 types of bounds. We demonstrate here how to replicate these using *gsDesign2*.
In *gsDesign2* the `gs_spending_bound()` function can be used to derive spending boundaries for all group sequential design derivations and power calculations. 
We demonstrate with the `gs_design_ahr()` function here, using designs under proportional hazards assumptions to compare with `gsDesign::gsSurv()`. 
Since the sample size methods differ between the `gsDesign2::gs_design_ahr()` and `gsDesign::gsSurv()` functions, we use continuous sample sizes so that spending bounds (Z-values, nominal p-values, spending) should be identical except where noted. 
Indeed, we are able to reproduce bounds to a high degree of accuracy.
Due to the different sample size methods, sample size and other boundary approximations vary slightly.

We also present a seventh example where to implement a futility bound based on observed hazard ratio as well as a Haybittle-Peto-like efficacy bound. In particular, the futility bound would be difficult to implement using the *gsDesign* package while it is straightforward using *gsDesign2*.

For all of our examples, we will use the following design assumptions:

```{r, message=FALSE}
library(gt)
library(gsDesign2)
library(gsDesign)
trial_duration <- 36         # Planned trial duration
info_frac <- c(.35, .7, 1)   # Information fraction at analyses
# 16 month planned enrollment with constant rate
enroll_rate <- define_enroll_rate(duration = 16, rate = 1)
# Minimum follow-up for gsSurv() (computed)
minfup <- trial_duration - enroll_rate$duration
# Failure rates
fail_rate <-
  define_fail_rate(
    duration = Inf,          # single time period, exponential failure
                             # 4 months and then remainder
    fail_rate = log(2) / 12, # Exponential time-to-event with 12 month median
    hr = .7,                 # Proportional hazards
    dropout_rate = -log(.99) / 12  # 1% dropout rate per year
  )
alpha <- 0.025  # Type I error
beta <- 0.15    # 85% power = 15% Type II error
ratio <- 1      # Randomization ratio
```



# Examples

Analogous to the *gsDesign* package, we look at 6 variations on combinations of efficacy and futility bounds.

## Example 1: Efficacy bound only

One-sided design has only an efficacy bound.
An easy way to do this is to use a fixed bound (`lower = gs_b`) with negative infinite bounds (`lpar = rep(-Inf, 3)`); in the table, infinite bounds do not appear.
The upper bound implements a spending bound (`upper = gs_spending_bound`) and the list of objects provided in `upar`.
The only parts of the list used here are `sf = gsDesign::sfLDOF` to select a Lan-DeMets spending function that approximates an O'Brien-Fleming bound. The `total_spend = alpha` sets the total spending to the targeted Type I error for the study.
The upper bound provides the Type I error control for the design as it is not specified elsewhere.


```{r}
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL, timing = NULL)

one_sided <- gsDesign2::gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_info", # Default
  # Upper spending bound and corresponding parameter(s)
  upper = gs_spending_bound, upar = upar,
  # No lower bound
  lower = gs_b, lpar = rep(-Inf, 3)
)

summary(one_sided) |> gsDesign2::as_gt(title = "Efficacy bound only", subtitle = "alpha-spending")
```

Check with `gsDesign::gsSurv().` As noted above, sample size and event counts vary slightly from the design derived using `gs_design_ahr()`.
This also results in different crossing probabilities under the alternate hypothesis at interim analyses.

```{r}
oneSided <-
   gsSurv(alpha = alpha, beta = beta, timing = info_frac, T = trial_duration, minfup = minfup,
          lambdaC = fail_rate$fail_rate, eta = fail_rate$dropout_rate, hr = fail_rate$hr,
          r = 32, tol = 1e-08, # Precision parameters for computations
          test.type = 1, # One-sided bound; efficacy only
          # Upper bound parameters
          sfu = upar$sf, sfupar = upar$param,
          )
oneSided |> gsBoundSummary()
```

Comparing Z-value bounds directly we see they are the same through approximately 6 digits with precision parameters chosen (`r=32, tol = 1e-08`):

```{r}
one_sided$bound$z - oneSided$upper$bound
```



## Example 2: Symmetric 2-sided design

We now derive a symmetric 2-sided design. 
This requires use of the argument `h1_spending = FALSE` to use $\alpha$-spending for both the upper and lower bounds.
While the lower bound is labeled as a futility bound in the table, it would be better termed an efficacy bound for control better than experimental treatment.

```{r}
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)
lpar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)

symmetric <- gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_h1_info", # Default
  # Function and parameter(s) for upper spending bound 
  upper = gs_spending_bound, upar = upar,
  lower = gs_spending_bound, lpar = lpar,
  # Symmetric design use binding bounds
  binding = TRUE,
  h1_spending = FALSE # Use null hypothesis spending
)

summary(symmetric) |> 
  gsDesign2::as_gt(title = "2-sided Symmetric Design", subtitle = "Single spending function")
```

We compare with `gsDesign::gsSurv()`.

```{r}
Symmetric <-
   gsSurv(test.type = 2, # Two-sided symmetric bound
          alpha = alpha, beta = beta, timing = info_frac, T = trial_duration, minfup=minfup, r = 32, tol = 1e-08,
          lambdaC = fail_rate$fail_rate, eta = fail_rate$dropout_rate, hr = fail_rate$hr,
          sfu = upar$sf, sfupar = upar$param
          ) 
Symmetric |> gsBoundSummary()
```

Comparing Z-value bounds directly, we again see approximately 6 digits of accuracy.

```{r}
filter(symmetric$bound, bound == 'upper')$z - Symmetric$upper$bound
filter(symmetric$bound, bound == 'lower')$z - Symmetric$lower$bound
```

## Example 3: Asymmetric 2-sided design with $\beta$-spending and binding futility

Designs with binding futility bounds are generally not considered acceptable for Phase 3 trials as Type I error is not controlled if a futility bound is crossed and the trial continues, a not infrequent occurrence.
For a Phase 2b study, this may be acceptable and results in a slightly smaller sample size than a comparable design with 
a non-binding futility  bound that is presented in Example 4.


```{r}
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)
lpar <- list(sf = gsDesign::sfHSD, total_spend = beta, param = -.5)

asymmetric_binding <- gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_info",
  # Function and parameter(s) for upper spending bound 
  upper = gs_spending_bound, upar = upar,
  lower = gs_spending_bound, lpar = lpar,
  # Asymmetric beta-spending design using binding bounds
  binding = TRUE,
  h1_spending = TRUE # Use beta-spending for futility
)

summary(asymmetric_binding) |>
  gsDesign2::as_gt(title = "2-sided asymmetric design with binding futility",
        subtitle = "Both alpha- and beta-spending used")
```

We compare with `gsDesign::gsSurv()`.

```{r}
asymmetricBinding <-
   gsSurv(test.type = 3, # Two-sided asymmetric bound, binding futility
          alpha = alpha, beta = beta, timing = info_frac, T = trial_duration, minfup = minfup,  r = 32, tol = 1e-07,
          lambdaC = fail_rate$fail_rate, eta = fail_rate$dropout_rate, hr = fail_rate$hr,
          sfu = upar$sf, sfupar = upar$param, sfl = lpar$sf, sflpar = lpar$param
          ) 
asymmetricBinding |> gsBoundSummary()
```

Comparing Z-value bounds directly, we again see approximately 6 digits of accuracy in spite of needing to relaxing accuracy to `tol = 1e-07` in the call to `gsSurv()` in order to get convergence.

```{r}
filter(asymmetric_binding$bound, bound == 'upper')$z - asymmetricBinding$upper$bound
filter(asymmetric_binding$bound, bound == 'lower')$z - asymmetricBinding$lower$bound
```


## Example 4: Aymmetric 2-sided design with $\beta$-spending and non-binding futility bound

In the *gsDesign* package, asymmetric designs with non-binding $\beta$-spending used for futility are the default design.
The objectives of this type of design include:

- Meaningful futility bounds to stop a trial early if no treatment benefit is emerging for the experimental treatment vs. control.
- Type I error is controlled even if the trial continues after a futility bound is crossed.


```{r}
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)
lpar <- list(sf = gsDesign::sfHSD, total_spend = beta, param = -.5)

asymmetric_nonbinding <- gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_info",
  # Function and parameter(s) for upper spending bound 
  upper = gs_spending_bound, upar = upar,
  lower = gs_spending_bound, lpar = lpar,
  # Asymmetric beta-spending design use binding bounds
  binding = FALSE,
  h1_spending = TRUE # Use beta-spending for futility
)

summary(asymmetric_nonbinding) |>
  gsDesign2::as_gt(title = "2-sided asymmetric design with non-binding futility",
        subtitle = "Both alpha- and beta-spending used")
```


We compare with `gsDesign::gsSurv()`.

```{r}
asymmetricNonBinding <-
   gsSurv(test.type = 4, # Two-sided asymmetric bound, non-binding futility
          alpha = alpha, beta = beta, timing = info_frac, T = trial_duration, minfup = minfup, r = 32, tol = 1e-08,
          lambdaC = fail_rate$fail_rate, eta = fail_rate$dropout_rate, hr = fail_rate$hr,
          sfu = upar$sf, sfupar = upar$param, sfl = lpar$sf, sflpar = lpar$param
          ) 
asymmetricNonBinding |> gsBoundSummary()
```

Comparing Z-value bounds directly, we again see approximately 6 digits of accuracy.

```{r}
filter(asymmetric_nonbinding$bound, bound == 'upper')$z - asymmetricNonBinding$upper$bound
filter(asymmetric_nonbinding$bound, bound == 'lower')$z - asymmetricNonBinding$lower$bound
```


## Example 5: Aymmetric 2-sided design with null hypothesis spending and binding futility bound

```{r}
alpha_star <- .5
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)
lpar <- list(sf = gsDesign::sfHSD, total_spend = alpha_star, param = 1)

asymmetric_safety_binding <- gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_info",
  # Function and parameter(s) for upper spending bound 
  upper = gs_spending_bound, upar = upar,
  lower = gs_spending_bound, lpar = lpar,
  # Asymmetric design use binding bounds
  binding = TRUE,
  h1_spending = FALSE # Use null-spending for futility
)

summary(asymmetric_safety_binding) |>
  gsDesign2::as_gt(title = "2-sided asymmetric safety design with binding futility",
        subtitle = "Alpha-spending used for both bounds, asymmetrically")
```


```{r}
asymmetricSafetyBinding <-
   gsSurv(test.type = 5, # Two-sided asymmetric bound, binding futility, H0 futility spending
          astar = alpha_star, # Total Type I error spend for futility
          alpha = alpha, beta = beta, timing = info_frac, T = trial_duration, minfup = minfup,
          lambdaC = fail_rate$fail_rate, eta = fail_rate$dropout_rate, hr = fail_rate$hr,
          sfu = upar$sf, sfupar = upar$param, sfl = lpar$sf, sflpar = lpar$param
          ) 
asymmetricSafetyBinding |> gsBoundSummary()
```

Comparing Z-value bounds directly, we again see approximately 6 digits of accuracy.
For `gsSurv()` this did not require the alternate arguments for `r` and `tol`.

```{r}
filter(asymmetric_safety_binding$bound, bound == 'upper')$z - asymmetricSafetyBinding$upper$bound
filter(asymmetric_safety_binding$bound, bound == 'lower')$z - asymmetricSafetyBinding$lower$bound
```

## Example 6: Aymmetric 2-sided design with null hypothesis spending and non-binding futility bound



```{r}
upar <- list(sf = gsDesign::sfLDOF, total_spend = alpha, param = NULL)
lpar <- list(sf = gsDesign::sfHSD, total_spend = alpha_star, param = 1)

asymmetric_safety_nonbinding <- gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_info",
  # Function and parameter(s) for upper spending bound 
  upper = gs_spending_bound, upar = upar,
  lower = gs_spending_bound, lpar = lpar,
  # Asymmetric design use non-binding bounds
  binding = FALSE,
  h1_spending = FALSE, # Use null-spending for futility
  test_lower = c(TRUE, TRUE, FALSE),
  test_upper = c(FALSE, TRUE, TRUE)
)

summary(asymmetric_safety_nonbinding) |>
  gsDesign2::as_gt(title = "2-sided asymmetric safety design with non-binding futility",
                   subtitle = "Alpha-spending used for both bounds, asymmetrically")
```

```{r}
asymmetricSafetyNonBinding <-
   gsSurv(test.type = 6, # Two-sided asymmetric bound, binding futility, H0 futility spending
          astar = alpha_star, # Total Type I error spend for futility
          alpha = alpha, beta = beta, timing = info_frac, T = trial_duration, minfup = minfup, r = 32, tol = 1e-08,
          lambdaC = fail_rate$fail_rate, eta = fail_rate$dropout_rate, hr = fail_rate$hr,
          sfu = upar$sf, sfupar = upar$param, sfl = lpar$sf, sflpar = lpar$param
          ) 
asymmetricSafetyBinding |> gsBoundSummary()
```

Comparing Z-value bounds directly, we again see approximately 6 digits of accuracy.

```{r}
filter(asymmetric_safety_nonbinding$bound, bound == 'upper')$z - asymmetricSafetyNonBinding$upper$bound
filter(asymmetric_safety_nonbinding$bound, bound == 'lower')$z - asymmetricSafetyNonBinding$lower$bound
```

## Example 7: Alternate bound types

We consider two types of alternative boundary computation approaches. 

- Computing futility bounds based on a hazard ratio.
- Computing efficacy bounds with a Haybittle-Peto or a related Fleming-Harrington-O'Brien approach.

We begin with a futility bound. We will consider a non-binding futility bound as it does not impact Type I error.
Assume the clinical trial team wishes to stop the trial at the first two interim analyses if a targeted interim hazard ratio is not achieved.
This approach can require a bit of iteration (trial and error) to incorporate the final design endpoint count; we skip over this iteration here.
We assume we wish to consider stopping for futility if a hazard ratio greater than 1 and 0.9 are observed at interim analyses 1 and 2 with 106 and 212 events observed, respectively. The final analysis is planned for 306 events.

```{r}
# Targeted events at interim and final analysis
# This is based on above designs and then adjusted, as necessary
targeted_events <- c(106, 212, 305)
```

We wish to translate the hazard ratios specified to corresponding Z-values; this can be done as follows.

```{r}
interim_futility_z <- -gsDesign::hrn2z(hr = c(1, .9), n = c(106, 212))
interim_futility_z
```
We will add a final futility bound of `-Inf`, indicating no final futility analysis; this gives us a vector of Z-value bounds for all analyses. 
For this type of bound, Type II error will be computed rather based on bounds rather than the spending approach were bounds are computed based on specified spending.

```{r}
lower <- gs_b; # Allows specifying fixed Z-values for futility
# Translated HR bounds to Z-value scale
lpar <- c(interim_futility_z, -Inf)
```

For the efficacy bound, we first consider a Haybittle-Peto fixed bound for interim analyses.

```{r}
upper = gs_b; upar = qnorm(c(.001, .001, .0023), lower.tail = FALSE)
```



and then use a spending approach suggested by @fleming1984designs where a fixed spend is considered for each analysis. 




```{r}
upper <- gs_b; upar <- qnorm(c(.001, .001, .0244), lower.tail = FALSE)
upper <- gs_spending_bound
upar <- list(sf = gsDesign::sfLinear, 
             total_spend = alpha, 
             param = c(c(106, 212) / 305, c(.001, .00165)/.025), 
             timing = NULL)

asymmetric_fixed_bounds <- gs_design_ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, beta = beta,
  # Information fraction at analyses and trial duration
  info_frac = info_frac, analysis_time = trial_duration,
  # precision parameters for computations
  r = 32, tol = 1e-08,
  # Use NULL information for Type I error, H1 information for Type II error
  info_scale = "h0_info",
  # Function and parameter(s) for upper spending bound 
  upper = upper, upar = upar,
  lower = lower, lpar = lpar,
  # Non-binding futility bounds
  binding = FALSE
) |> to_integer()

summary(asymmetric_fixed_bounds) |>
  gsDesign2::as_gt(title = "2-sided asymmetric safety design with fixed non-binding futility",
                   subtitle = "Futility bounds computed to approximate HR")
```

# References

