---
title: "fixed_design_*: compute sample size/power of a fixed design"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    highlight: "textmate"
    css: "custom.css"
bibliography: gsDesign2.bib
vignette: |
  %\VignetteIndexEntry{fixed_design_*: compute sample size/power of a fixed design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, message=FALSE, warning=FALSE}
library(gt)
library(gsDesign)
library(gsDesign2)
library(dplyr)
```

# Introduction

In this vignette, we introduce the key functionality of `fixed_design_xxx()` in `gsDesign2`.
Generally speaking, `fixed_design_xxx()` is capable of both power calculations and sample size calculations under many methods, including

- `fixed_design_ahr()`: AHR
- `fixed_design_fh()`Fleming-Harrington (FH) [@farrington1990test]
- `fixed_design_mb()`Magirr-Burman (MB)
- `fixed_design_lf()`Lachin-Foulkes (LF) [@lachin1986evaluation]
- `fixed_design_rmst()`RMST [@yung2020sample]
- `fixed_design_milestone()`Milestone [@yung2020sample]
- `fixed_design_rd()`Risk difference (RD) [@mehrotra2000minimum].

Throughout this vignette, we assume the design parameter is

```{r}
# Enrollment rate
enroll_rate <- define_enroll_rate(
  duration = 18,
  rate = 20
)

# Failure rates
fail_rate <- define_fail_rate(
  duration = c(4, 100),
  fail_rate = log(2) / 12,
  hr = c(1, .6),
  dropout_rate = .001
)

# Study duration in months
study_duration <- 36

# Experimental / Control randomization ratio
ratio <- 1

# 1-sided Type I error
alpha <- 0.025
# Type II error (1 - power)
beta <- 0.1
```

# Examples of AHR {.tabset}

If one wants to get a power of 90%, one can calculate the sample size by

```{r}
fixed_design_ahr(
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of AHR by

```{r}
fixed_design_ahr(
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio
) %>%
  summary() %>%
  as_gt()
```

# Examples of FH {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
# Fixed design with a given power with default rho/gamma
fixed_design_fh(
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio
) %>%
  summary() %>%
  as_gt()
```

In the above example, the FH parameters (`rho`, `gamma`) take the default value as `rho = 0`, `gamma = 0.5`. But users can always custom their `rho` and `gamma` by

```{r}
# Fixed design with a given power with input rho/gamma
fixed_design_fh(
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio,
  rho = 0.5,
  gamma = 0.5
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of FH under the default `rho`, `gamma` by

```{r}
# Fixed design with power calculated
fixed_design_fh(
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio
) %>%
  summary() %>%
  as_gt()
```

But users can always custom their `rho` and `gamma` by

```{r}
# Fixed design with power calculated
fixed_design_fh(
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio,
  rho = 0.5, gamma = 0.5
) %>%
  summary() %>%
  as_gt()
```

# Examples of mb {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
fixed_design_mb(
  ratio = ratio,
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
) %>%
  summary() %>%
  as_gt()
```

In the above example, the mb parameter (`tau`) takes the default value as `tau = 6`. But users can always custom their `tau` by

```{r}
fixed_design_mb(
  ratio = ratio,
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  tau = 4
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of MB under the default `tau = 6` by

```{r}
fixed_design_mb(
  ratio = ratio,
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
) %>%
  summary() %>%
  as_gt()
```

But users can always custom their `tau` by

```{r}
fixed_design_mb(
  ratio = ratio,
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  tau = 4
) %>%
  summary() %>%
  as_gt()
```

# Examples of lf {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
fixed_design_lf(
  alpha = alpha,
  power = 1 - beta,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of LF by

```{r}
fixed_design_lf(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
) %>%
  summary() %>%
  as_gt()
```

# Examples of MaxCombo {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
fixed_design_maxcombo(
  alpha = alpha,
  power = 1 - beta,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
) %>%
  summary() %>%
  as_gt()
```

In the above example, the MaxCombo parameters take the default value

```{r}
data.frame(rho = c(0, 0), gamma = c(0, 0.5), tau = -1) %>% gt()
```

But users can always custom their `rho`, `gamma`, `tau` by

```{r}
fixed_design_maxcombo(
  alpha = alpha,
  power = 1 - beta,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  rho = c(0, 0.5, 0.5),
  gamma = c(0, 0, 0.5),
  tau = c(-1, 4, 6)
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of MaxCombo under the default parameters by

```{r}
fixed_design_maxcombo(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
) %>%
  summary() %>%
  as_gt()
```

But users can always custom their `rho`, `gamma`, `tau` by

```{r}
fixed_design_maxcombo(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  rho = c(0, 0.5, 0.5),
  gamma = c(0, 0, 0.5),
  tau = c(-1, 4, 6)
) %>%
  summary() %>%
  as_gt()
```

# Examples of RMST {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
fixed_design_rmst(
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio,
  tau = 18
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of RMST by

```{r}
fixed_design_rmst(
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio,
  tau = 18
) %>%
  summary() %>%
  as_gt()
```

# Examples of Milestone {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
fixed_design_milestone(
  alpha = alpha,
  power = 1 - beta,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio,
  tau = 18
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power of Milestone by

```{r}
fixed_design_milestone(
  alpha = alpha,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  ratio = ratio,
  tau = 18
) %>%
  summary() %>%
  as_gt()
```

# Examples of RD {.tabset}

If one wants to get a power of 90\%, one can calculate the sample size by

```{r}
fixed_design_rd(
  alpha = alpha,
  power = 1 - beta,
  p_c = .15,
  p_e = .1,
  rd0 = 0,
  ratio = ratio
) %>%
  summary() %>%
  as_gt()
```

Given the above enrollment rate and enrollment duration, one can calculate the power by

```{r}
fixed_design_rd(
  alpha = alpha,
  power = NULL,
  p_c = .15,
  p_e = .1,
  rd0 = 0,
  n = 2000,
  ratio = ratio
) %>%
  summary() %>%
  as_gt()
```

# Examples to get multiple designs together

```{r, message=FALSE}
x_ahr <- fixed_design_ahr(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
)

x_fh <- fixed_design_fh(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  rho = 0.5,
  gamma = 0.5
)

x_mb <- fixed_design_mb(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  tau = 4
)

x_lf <- fixed_design_lf(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration
)

x_maxcombo <- fixed_design_maxcombo(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  rho = c(0, 0.5, 0.5),
  gamma = c(0, 0, 0.5),
  tau = c(-1, 4, 6)
)

x_rmst <- fixed_design_rmst(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  tau = 30
)

x_milestone <- fixed_design_milestone(
  alpha = alpha,
  ratio = ratio,
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  study_duration = study_duration,
  tau = 30
)

rbind(
  summary(x_ahr), summary(x_fh),
  summary(x_mb), summary(x_lf),
  summary(x_maxcombo), summary(x_rmst),
  summary(x_milestone)
) %>% gt()
```
