---
title: "expected_event: compute expected number of events at 1 time point"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{expected_event: compute expected number of events at 1 time point}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.align = "center")
```

```{r, message=FALSE, warning=FALSE}
library(gt)
library(dplyr)
library(testthat)
library(gsDesign2)
```

# Introduction of `expected_event()`

`expected_event()` computes expected number of events at a **given analysis time by strata** under the assumption of piecewise model:

- piecewise constant enrollment rates
- piecewise exponential failure rates
- piecewise censoring rates.

The above piecewise exponential distribution allows a simple method to specify a distribution and enrollment pattern where the enrollment, failure and dropout rates changes over time.

Here the `df` in `expected_event()` is short for data frame, since its output is a data frame.

# Use Cases

## Example 1: Single Enroll + Single Fail Period

```{r}
enroll_rate <- define_enroll_rate(duration = 10, rate = 10)
fail_rate <- define_fail_rate(duration = 100, fail_rate = log(2) / 6, dropout_rate = .01)
total_duration <- 22

expected_event(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  total_duration = total_duration,
  simple = FALSE
)
```

## Example 2: Multiple Enroll + Single Fail Period

```{r}
enroll_rate <- define_enroll_rate(duration = c(5, 5), rate = c(10, 20))
fail_rate <- define_fail_rate(duration = 100, fail_rate = log(2) / 6, dropout_rate = .01)
total_duration <- 22

expected_event(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  total_duration = total_duration,
  simple = FALSE
)
```

## Example 3: Single Enroll + Multiple Fail Period

```{r}
enroll_rate <- define_enroll_rate(duration = 10, rate = 10)
fail_rate <- define_fail_rate(
  duration = c(20, 80),
  fail_rate = c(log(2) / 6, log(2) / 4),
  dropout_rate = .01
)
total_duration <- 22

expected_event(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  total_duration = total_duration,
  simple = FALSE
)
```

# Inner Logic of `expected_event()`

**Step 1:** set the analysis time.

```{r}
total_duration <- 50
```

**Step 2:** set the enrollment rates.

```{r}
enroll_rate <- define_enroll_rate(duration = c(5, 5), rate = c(10, 20))

# create a step function (sf) to define enrollment rates over time
sf_enroll_rate <- stepfun(c(0, cumsum(enroll_rate$duration)),
  c(0, enroll_rate$rate, 0),
  right = FALSE
)

plot(sf_enroll_rate,
  xlab = "duration", ylab = "enrollment rates",
  main = "Piecewise enrollment rate over time", xlim = c(-0.01, 21)
)
```

**Step 3:** set the failure rates and dropout rates.

```{r}
fail_rate <- define_fail_rate(
  duration = c(20, 80),
  fail_rate = c(0.1, 0.2),
  dropout_rate = .01
)

# get the time points where the failure rates change
time_start_fail <- c(0, cumsum(fail_rate$duration))

# plot the piecewise failure rates
sf_fail_rate <- stepfun(time_start_fail,
  c(0, fail_rate$fail_rate, last(fail_rate$fail_rate)),
  right = FALSE
)
plot(sf_fail_rate,
  xlab = "duration", ylab = "failure rates",
  main = "Piecewise failure rate over time", xlim = c(-0.01, 101)
)

# plot the piecewise dropout rate
sf_dropout_rate <- stepfun(time_start_fail,
  c(0, fail_rate$dropout_rate, last(fail_rate$dropout_rate)),
  right = FALSE
)
plot(sf_dropout_rate,
  xlab = "duration", ylab = "dropout rates",
  main = "Piecewise dropout rate over time", xlim = c(-0.01, 101)
)
```

Given the above piecewise enrollment rates, failure rates, dropout rates, the time line is divided into several parts:

- $(0, 5]$ (5 is the change point of the enrollment rates);
- $(5, 10]$ (10 is another change point of the enrollment rates);
- $(10, 20]$ (20 is the change point of the failure rates);
- $(20, 50]$ (50 is the analysis time);
- $(50, \infty]$ (after the analysis time).

```{r, echo=FALSE}
plot(sf_enroll_rate,
  xlab = "time", ylab = "enrollment rates",
  lty = 1, col = "red", pch = 1,
  main = "Piecewise enrollment, failure, dropout rate over time", xlim = c(-0.01, 101)
)

plot(sf_fail_rate,
  xlab = "time", ylab = "failure rates",
  lty = 2, col = "blue", pch = 2,
  xlim = c(-0.01, 101), add = TRUE
)

plot(sf_dropout_rate,
  xlab = "time", ylab = "dropout rates",
  lty = 3, col = "green", pch = 3,
  xlim = c(-0.01, 101), add = TRUE
)

legend(60, 20, c("enrollment rate", "failure rate", "dropout rate"),
  col = c("red", "blue", "green"),
  lty = c(1, 2, 3), pch = c(1, 2, 3),
  merge = TRUE, bg = "gray90"
)
```

Given the above sub-intervals, our objective is to calculate the expected number of events in each sub-intervals.

**Step 4:** divide the time line for enrollments

```{r}
df_1 <- tibble(
  startEnroll = c(0, cumsum(enroll_rate$duration)),
  endFail = total_duration - startEnroll,
  rate = c(enroll_rate$rate, 0)
)
```

```{r, echo=FALSE}
df_1 %>%
  gt() %>%
  tab_header(title = "df_1") %>%
  tab_footnote(
    footnote = "The time when the enrollment starts.",
    locations = cells_column_labels("startEnroll")
  ) %>%
  tab_footnote(
    footnote = "The time from startEnroll to the analysis time.",
    locations = cells_column_labels("endFail")
  ) %>%
  tab_footnote(
    footnote = "The enrollment rates",
    locations = cells_column_labels("rate")
  )
```

**Step 5:** divide the time line for failure \& dropout rates

```{r}
df_2 <- tibble(
  endFail = cumsum(fail_rate$duration),
  startEnroll = total_duration - endFail,
  failRate = fail_rate$fail_rate,
  dropoutRate = fail_rate$dropout_rate
)
```

```{r, echo=FALSE}
df_2 %>%
  gt() %>%
  tab_header(title = "df_2") %>%
  tab_footnote(
    footnote = "The time when the failure changes.",
    locations = cells_column_labels("endFail")
  ) %>%
  tab_footnote(
    footnote = "The time from endFail to the analysis time.",
    locations = cells_column_labels("startEnroll")
  ) %>%
  tab_footnote(
    footnote = "The failure rates",
    locations = cells_column_labels("failRate")
  ) %>%
  tab_footnote(
    footnote = "The dropout rates",
    locations = cells_column_labels("dropoutRate")
  )
```

For the above `df_2`, one needs to discriminate if the analysis time (`total_duration = 50`) is beyond the total failure rate duration.

```{r}
# if the analysis time is after the total failure rate duration
if (sum(fail_rate$duration) < total_duration) {
  df_2 <- df_2[-nrow(df_2), ]
} else {
  df_2 <- df_2 %>% filter(startEnroll > 0)
}
```

```{r, echo=FALSE}
df_2 %>%
  gt() %>%
  tab_header(
    title = "df_2",
    subtitle = "Updated by adjusting the analysis time and fail_rate duration"
  ) %>%
  tab_footnote(
    footnote = "The time when the failure changes.",
    locations = cells_column_labels("endFail")
  ) %>%
  tab_footnote(
    footnote = "The time from endFail to the analysis time.",
    locations = cells_column_labels("startEnroll")
  ) %>%
  tab_footnote(
    footnote = "The failure rates",
    locations = cells_column_labels("failRate")
  ) %>%
  tab_footnote(
    footnote = "The dropout rates",
    locations = cells_column_labels("dropoutRate")
  )
```

**Step 6:** divide the time line considering both the change points in enrollment, failure, dropout rates.

```{r}
df <- full_join(df_1, df_2, by = c("startEnroll", "endFail")) %>% arrange(endFail)
```

```{r, echo=FALSE}
df %>%
  gt() %>%
  tab_header(title = "df") %>%
  tab_footnote(
    footnote = "The time when the enrollment rate starts.",
    locations = cells_column_labels("startEnroll")
  ) %>%
  tab_footnote(
    footnote = "The time when the failure rate ends. And startEnroll + endFail = 50",
    locations = cells_column_labels("endFail")
  ) %>%
  tab_footnote(
    footnote = "The enrollment rates.",
    locations = cells_column_labels("rate")
  )
```

We find there are lots of `NA`, which can be imputed by the piecewise model.
```{r}
df <- df %>%
  mutate(
    endEnroll = lag(startEnroll, default = as.numeric(total_duration)),
    time_start_fail = lag(endFail, default = 0),
    duration = endEnroll - startEnroll,
    failRate = sf_fail_rate(time_start_fail),
    dropoutRate = sf_dropout_rate(time_start_fail),
    enrollRate = sf_enroll_rate(startEnroll)
  ) %>%
  select(-rate)
```

```{r, echo=FALSE}
df %>%
  select(startEnroll, endEnroll, time_start_fail, endFail, enrollRate, failRate, dropoutRate, duration) %>%
  arrange(startEnroll) %>%
  gt() %>%
  tab_footnote(
    footnote = "The time when the enrollment rate starts.",
    locations = cells_column_labels("startEnroll")
  ) %>%
  tab_footnote(
    footnote = "The (startEnroll, endEnroll] forms the piecewise model of the enrollment rates",
    locations = cells_column_labels("endEnroll")
  ) %>%
  tab_footnote(
    footnote = "The time when the failure rate starts.",
    locations = cells_column_labels("time_start_fail")
  ) %>%
  tab_footnote(
    footnote = "The time when the failure rate ends.
    And startEnroll + endFail = 50.
    Besides, (time_start_fail, endFail  ] forms
    the piecewise model of the enrollment rates.",
    locations = cells_column_labels("endFail")
  ) %>%
  tab_footnote(
    footnote = "endEnroll - startEnroll",
    locations = cells_column_labels("duration")
  )
```

**Step 7:** compute the expected number of events in sub-intervals following the technical details in the vignette [computing expected events by interval at risk](https://merck.github.io/gsDesign2/articles/eEventsTheory.html).

```{r}
# Create 2 auxiliary variables for failure & dropout rate
# q: number of expected events in a sub-interval
# Q: cumulative product of q (pool all sub-intervals)
df <- df %>%
  mutate(
    q = exp(-duration * (failRate + dropoutRate)),
    Q = lag(cumprod(q), default = 1)
  ) %>%
  arrange(desc(time_start_fail)) %>%
  # Create another 2 auxiliary variables for enroll rate
  # g: number of expected subjects in a sub-interval
  # G: cumulative sum of g (pool all sub-intervals)
  mutate(
    g = enrollRate * duration,
    G = lag(cumsum(g), default = 0)
  ) %>%
  arrange(time_start_fail) %>%
  # Compute expected events as nbar in a sub-interval
  mutate(
    d = ifelse(
      failRate == 0,
      0,
      Q * (1 - q) * failRate / (failRate + dropoutRate)
    ),
    nbar = ifelse(
      failRate == 0,
      0,
      G * d + (failRate * Q * enrollRate) / (failRate + dropoutRate) * (duration - (1 - q) / (failRate + dropoutRate))
    )
  )
```

**Step 8:**  output results

```{r}
sf_time_start_fail <- stepfun(time_start_fail, c(0, time_start_fail), right = FALSE)
df <- df %>%
  transmute(t = endFail, failRate = failRate, Events = nbar, time_start_fail = sf_time_start_fail(time_start_fail)) %>%
  group_by(time_start_fail) %>%
  summarize(failRate = first(failRate), Events = sum(Events)) %>%
  mutate(t = time_start_fail) %>%
  select("t", "failRate", "Events")

df %>% gt()
```
