---
title: "expected_time: compute time when a targeted number of events is made"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{expected_time: compute time when a targeted number of events is made}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.align = "center")
```

```{r, message=FALSE, warning=FALSE}
library(gt)
library(tibble)
library(dplyr)
library(testthat)
library(gsDesign2)
```

# Introduction of `expected_time`

`expected_time()` predicts time at which a targeted events is made.
It is designed as a twins to `AHR()`: it matches input/output format with `AHR()`.

# Use Cases

## Example 1

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18), hr = c(.9, .6),
  dropout_rate = rep(.001, 2)
)
ratio <- 1

x <- expected_time(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, target_event = 200
)

x %>% gt()
```

## Example 2

In this example, we verify `expected_time()` by `AHR()`.

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = rep(.001, 2)
)
ratio <- 1

x <- ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, total_duration = 20
)
cat("The number of events by 20 months is ", x$event, ".\n")

y <- expected_time(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  ratio = ratio, target_event = x$event
)

cat("The time to get ", x$event, " is ", y$time, "months.\n")
```

# Inner Logic of `expected_time()`

The inner logic of `expected_time()` is to uniroot `AHR()` on `total_duration`.

**Step 1:** find the difference between  `AHR()` and different values of `total_duration`.

```{r}
foo <- function(x) {
  ans <- ahr(
    enroll_rate = enroll_rate, fail_rate = fail_rate,
    total_duration = x, ratio = ratio
  )$event - target_event
  return(ans)
}
```

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 2, 10),
  rate = c(3, 6, 9) * 5
)
fail_rate <- define_fail_rate(
  duration = c(3, 100),
  fail_rate = log(2) / c(9, 18),
  hr = c(.9, .6),
  dropout_rate = .001
)
ratio <- 1
target_event <- 200

cat("The difference between `targetEvents = 200` and the events after 30 months is ", foo(30), ".\n")
```

**Step 2:** uniroot `AHR()` on `total_duration`.

```{r}
res <- uniroot(foo, interval = c(0.01, 100))

ans <- ahr(
  enroll_rate = enroll_rate, fail_rate = fail_rate,
  total_duration = res$root, ratio = ratio
)
cat("After ", ans$time, " months, there will be ", target_event, " events .\n")
```
