---
title: "ahr: computes AHR under NPH assumptions and (stratified) populations"
author: "Yujie Zhao"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    highlight: "textmate"
    css: "custom.css"
# bibliography: "example.bib"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ahr: computes AHR under NPH assumptions and (stratified) populations}
---

```{r, message=FALSE, warning=FALSE}
library(tibble)
library(dplyr)
library(gsDesign2)
```

# Introduction of `ahr()`

`ahr()` provides a **geometric average hazard ratio** under various non-proportional hazards assumptions for either single or multiple strata studies.
The piecewise exponential distribution allows a simple method to specify a distribution and enrollment pattern where the enrollment, failure and dropout rates changes over time.

# Usage of `ahr()`

## Example 1: Un-stratified population

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 10, 4, 4, 8),
  rate = c(5, 10, 0, 3, 6)
)
fail_rate <- define_fail_rate(
  duration = 1,
  fail_rate = c(.1, .2, .3, .4),
  hr = c(.9, .75, .8, .6),
  dropout_rate = .001
)
ahr(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  total_duration = c(15, 30)
) %>% gt::gt()
```

## Example 2: Stratified population

```{r}
enroll_rate <- define_enroll_rate(
  stratum = c(rep("Low", 2), rep("High", 3)),
  duration = c(2, 10, 4, 4, 8),
  rate = c(5, 10, 0, 3, 6)
)
fail_rate <- define_fail_rate(
  stratum = c(rep("Low", 2), rep("High", 2)),
  duration = 1,
  fail_rate = c(.1, .2, .3, .4),
  hr = c(.9, .75, .8, .6),
  dropout_rate = .001
)
ahr(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate,
  total_duration = c(15, 30)
) %>% gt::gt()
```

# Inner Logic of `ahr()`

Let's take the un-stratified population as an example, where the enrollment rate, failure rates and dropout rates are

```{r}
enroll_rate <- define_enroll_rate(
  duration = c(2, 10, 4),
  rate = c(5, 10, 0)
)
fail_rate <- define_fail_rate(
  duration = 1,
  fail_rate = c(.1, .2),
  hr = c(.9, .75),
  dropout_rate = .001
)

ratio <- 2

total_duration <- 30
```

**Step 1:** compute proportion in each group

```{r}
q_e <- ratio / (1 + ratio)
q_c <- 1 - q_e
```

```{r, echo=FALSE}
cat("The proportion of the experimental arm is ", q_e, "\n")
cat("The proportion of the control arm is ", q_c, "\n")
```

To compute the expected events over different treatment group, stratum and time period, we iterate over `total_duration` and `Strata`.
Since in this example, we only have one analysis time (`total_duration = 30`) and one stratum (`stratum = "All"`), we only iterate once.
In one has multiple analysis time and strata, one can use a for loop and bind the results by row.

```{r}
td <- total_duration
s <- "All"
```

**Step 2:** subset the enrollment rates and failure rates by stratum.

```{r}
enroll <- enroll_rate %>% filter(stratum == s)
fail <- fail_rate %>% filter(stratum == s)
```

**Step 3:** we calculate the enrollment rates in experimental arm and control arm, respectively.

```{r}
enroll_c <- enroll %>% mutate(rate = rate * q_c)
enroll_e <- enroll %>% mutate(rate = rate * q_e)
```

**Step 4:** we update the failure rate in the control and experimental arm.

```{r}
fail_c <- fail
fail_e <- fail %>% mutate(fail_rate = fail_rate * hr)
```

**Step 5:** we calculate the expected number of events in the control and experimental by `eEvents_df()`.

```{r}
events_c <- expected_event(
  enroll_rate = enroll_c,
  fail_rate = fail_c,
  total_duration = td,
  simple = FALSE
)
events_e <- expected_event(
  enroll_rate = enroll_e,
  fail_rate = fail_e,
  total_duration = td,
  simple = FALSE
)
```

```{r, echo=FALSE}
cat("The expected number of events in the control arm is \n")
events_c

cat("The expected number of events in the experimental arm is \n")
events_e
```

Here the `t` column is the start of period, the `fail_rate` column is the failure rate during the period, and the `Events` column is the  expected events during the period.

**Step 6:** we combine the results together and output it.

```{r, message=FALSE}
# combine control and experimental
events <- rbind(
  events_c %>% mutate(Treatment = "Control"),
  events_e %>% mutate(Treatment = "Experimental")
) %>%
  arrange(t, Treatment) %>%
  ungroup() %>%
  # recompute HR, events, info by period
  group_by(t) %>%
  summarize(
    stratum = s,
    info = (sum(1 / event))^(-1),
    Events = sum(event),
    HR = last(fail_rate) / first(fail_rate)
  ) %>%
  # compute info0
  mutate(
    Time = td,
    lnhr = log(HR),
    info0 = Events * q_c * q_e
  ) %>%
  ungroup() %>%
  group_by(Time, stratum, HR) %>%
  summarize(
    t = min(t),
    Events = sum(Events),
    info0 = sum(info0),
    info = sum(info)
  ) %>%
  # pool time period together
  group_by(Time) %>%
  summarize(
    AHR = exp(sum(log(HR) * Events) / sum(Events)),
    Events = sum(Events),
    info = sum(info),
    info0 = sum(info0)
  )
```

```{r, echo=FALSE}
cat("The overall expected number of events over the time is \n")
events
```

Please note that, in the output, the `info` column is based on the following input.

```{r, eval=FALSE}
enroll_rate <- define_enroll_rate(
  duration = c(2, 10, 4),
  rate = c(5, 10, 0)
)
fail_rate <- define_fail_rate(
  duration = 1,
  fail_rate = c(.1, .2),
  hr = c(.9, .75),
  dropout_rate = .001
)
```

If the alternative hypothesis $H_1$ is

$$
  \text{hr}
  =
  \left\{
  \begin{array}{ll}
    0.9  & \text{for the first 1 month} \\
    0.75 & \text{afterwards},
  \end{array}
  \right.
$$

then `info = info1`, where `info1` is the statistical information under $H_1$.
But notice that the above `enroll_rate` and `fail_rate` is not always the $H_1$,
so we call it as `info`, rather than `info1`.
